### 2024-11-26 paper/paper.qmd
Update complete paper

---
title: "Environmental Conditions and Driver Behavior Shape Collision Outcomes in Toronto"
subtitle: "A Two-Stage Bayesian Analysis Reveals Infrastructure Factors Predict Crash Occurrence While Driver Actions Determine Severity"
author: Yitong Wang
thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
date: today
date-format: long
abstract: "This study employs a two stage bayesian modeling approach to analyze vehicle collision patterns in Toronto, examining how environmental conditions and driver behavior influence both injury occurrence and severity. The first model predicts the probability of injury based on environmental and infrastructural factors, while the second estimates injury severity using driver characteristics and behavior. Analysis of historical collision data from 2006 to 2023 reveals that while environmental conditions primarily influence injury occurrence, driver behavior factors like speeding and alcohol use more strongly affect severity levels. Our dual model analysis providing evidence based guidance for targeting both infrastructure improvements and enforcement strategies to reduce severe collision outcomes."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
library(tidyverse)
library(ggplot2)
library(dplyr)
library(pROC)
library(brms)
library(arrow)
library(here)
library(bayesplot)
library(knitr)
library(modelsummary)

cleaned_data <- read_parquet(here("data", "02-analysis_data", "analysis_data.parquet"))
occurrence_model <- read_rds(here("models", "occurrence_model.rds"))
severity_model <- read_rds(here("models", "severity_model.rds"))
severity_test <- read_parquet(here("data", "03-model_related_data", "severity_test.parquet"))
```

# Introduction
Vehicle collisions in urban environments represent a significant public health and safety challenge, with particular concern in large metropolitan areas like Toronto where complex traffic patterns and diverse environmental conditions create varying levels of risk. Understanding the factors that contribute to collision occurrence and severity is crucial for urban planning and public safety policy, yet traditional analyses often treat these as a single outcome rather than recognizing their distinct underlying processes. The increasing availability of detailed collision data, combined with advances in statistical modeling, presents an opportunity to develop more nuanced understandings of how different factors influence collision outcomes.

This paper aims to address this challenge through a two-stage modeling approach, with the estimand being the predicted probability of injury occurrence and, conditional on an injury occurring, its severity level. Specifically, our first model examines how environmental and infrastructural factors influence the likelihood of injury in a collision, while the second model investigates how driver characteristics and behavior affect injury severity when injuries do occur. By separating these processes, we can better understand how different factors contribute to collision outcomes and identify more targeted intervention opportunities.

Our analysis reveals several distinct and important patterns in how different factors influence collision outcomes. Environmental conditions emerged as the dominant predictors of whether an injury occurs in the first place - particularly reduced visibility during winter months and poor road surface conditions on major arterials. However, once an injury occurs, driver behavior becomes the crucial determinant of severity. Impaired driving and speeding showed strong associations with severe outcomes, especially during nighttime hours and weekends. Road classification provided an interesting counterintuitive finding: while major arterial roads saw more frequent injuries, the injuries on local roads tended to be more severe. Temporal analysis uncovered clear risk patterns, with injury probability peaking during adverse weather conditions on arterials, while severity levels were highest during weekend nights with notable correlations to driver behavior violations. Pedestrian and cyclist collisions showed a particularly concerning pattern, even overall less frequent, they consistently resulted in more severe injuries across all conditions.

By separating environmental triggers from behavioral factors, we uncovered distinct patterns that shape Toronto's traffic safety landscape. Our analysis shows how road design and maintenance decisions directly influence where injuries occur, while enforcement timing determines crash severity. Understanding these relationships helps transportation departments focus their resources where they matter most, from targeted winter maintenance on high risk arterials to enforcement during periods when severe crashes are most likely. Beyond potential safety improvements, this research advances our understanding of urban traffic patterns and provides a foundation for evidence-based policy decisions.

The remainder of this paper is structured as follows:
@sec-data presents the overview of data (@sec-data-overview), data cleaning process (@sec-data-cleaning), measurement (@sec-data-measure), also for explanations, descriptions, graph summaries of outcome (@sec-data-outcome) and predictor variables (@sec-data-predictor) of the study. @sec-model set up the model and explain the process detailed, including procedure of model set up (@sec-model-setup) and model validation (@sec-model-validate), and also mention the assumption and validation (@sec-model-assumpandlimit) of models. In addition, (@sec-model-alternate) discuss the potential alternative models and its trade offs. Then, @sec-result reveal the prediction outcome and models' performance. @sec-discussion discusses the relation between the data and real world, which talks about the data quality and bias, as well as the potential limitation of our models in a broader context.

Appendix includes three parts: [Appendix -@sec-data-methodology] presents collision data methodology overview, [Appendix -@sec-data-idealized] provides detailed idealized methodology for collection improvement; [Appendix -@sec-model-details] presents additional summary and validation in modeling process.

# Data {#sec-data}
## Overview {#sec-data-overview}
The data for this analysis comes from Toronto's Motor Vehicle Collision dataset [@toronto_collision_data], accessed through the Toronto Open Data [@opendatatoronto]. The raw dataset contains 14,976 records of vehicle collisions involving killed or different levels injured persons from 2006 to 2023, with 52 variables capturing various aspects of each incident. These include collision characteristics, environmental conditions, road features, and driver factors.

In this study, we utilized statistical language R [@citeR] and several key packages for data processing, analysis, and visualization. The tidyverse package [@tidyverse] facilitated data manipulation, with dplyr [@dplyr] being particularly crucial for data transformation and filtering operations, while visualization was conducted using ggplot2 [@ggplot2]. For efficient and consistent data import, we employed the readr package [@readr]. For handling temporal aspects of collision data, we employes lubridate {@lubridate} to process dates and times.We used here [@here] for reproducible file path management, arrow [@arrow] for efficient data storage and retrieval, and knitr [@knitr] used to provide the better formatting output. For unit testing and code validation, we employed the testthat package [@testthat] which provided a robust framework for testing our functions and ensuring code reliability. Brms package [@brms] plays the significant values for our model construction. 
For Bayesian analysis visualization, we employed bayesplot [@bayesplot] to create diagnostic plots including MCMC trace plots. For model evaluation, we employed the pROC package [@pROC] to generate receiver operating characteristic ROC curves and assess model performance. For model visualization and interpretation, we used the modelsummary package [@modelsummary] to create coefficient plots and model summaries.

## Data Cleaning {#sec-data-cleaning}
From the raw dataset, we selected 14 key variables most relevant to understanding collision severity patterns: date, time, road_class (road classification), accloc (accident location), traffctl (traffic control), visibility, light, rdsfcond (road surface condition), impactype (impact type), invage (involved person's age), injury (severity), speeding, ag_driv (aggressive driving), alcohol, and disability. Records with missing values in critical fields such as injury severity, visibility, or road classification were removed to ensure data quality.

In addition to removing rows with missing NA values, we also excluded records where key categorical variables were explicitly recorded as "None" despite an incident occurring. This includes records where accident location (accloc), road classification (road_class), traffic control (traffctl), visibility, light conditions (light), road surface condition (rdsfcond), and impact type (impactype) were recorded as "None". These "None" entries differ from missing values as they were actively recorded but indicate potentially incomplete or unclear documentation of the collision circumstances. This additional cleaning step ensures our analysis is based on collisions with well-documented characteristics, though it reduced our dataset by approximately 8% (1,057 records). This filtering helps maintain data quality and supports more reliable analysis of factor relationships with collision severity.

## Measurement {#sec-data-measure}
The transformation of real-world collision incidents into structured data points follows a rigorous protocol established by Toronto Police Services. When a collision occurs, first responders and investigating officers collect data through a standardized Motor Vehicle Accident Report Form. This form ensures systematic documentation of collision characteristics, converting complex real-world events into quantifiable measurements. For instance, injury severity classification follows specific criteria: fatalities are recorded when death occurs within 30 days of collision, major injuries are those requiring hospitalization, minor injuries need medical attention but no hospitalization, and minimal injuries indicate no significant medical intervention was required. This standardized approach helps maintain consistency in severity classification across different incidents and responding officers [@toronto_police_services_ksi_guide].

Environmental and situational measurements are recorded at the time of collision through both objective and subjective assessments. Objective measurements include temporal factors (date, time), location (coordinates, road type), and certain environmental conditions (light, road conditions). However, some measurements inherently involve officer judgment, such as assessing visibility conditions (Clear, Rain, Snow) or driver behavior (aggressive driving, speeding). These assessments, while guided by training and protocols, may introduce some variability in how similar conditions are classified across different incidents [@mvcr_guide]. For example, what one officer considers "aggressive driving" might be classified differently by another, highlighting the importance of considering potential measurement inconsistencies in our analysis.

Some driver related factors present particular measurement challenges due to their complex nature. Alcohol involvement is measured through standardized tests when possible, and it providing more objective data. However, other driver conditions like fatigue or distraction often rely on post incident interviews and officer's observation, which introducing potential reporting biases. Additionally, factors like driver age and license status are verified through official documentation, ensuring accuracy in demographic data. The measurement process also captures infrastructure characteristics such as road classification and traffic control presence, which are determined through existing municipal records and on-site verification. This multi-faceted approach to data collection, while comprehensive, means that the quality and reliability of different variables may vary depending on the objectivity of their measurement methods.

## Outcome variables {#sec-data-outcome}
Our primary outcome variable is injury severity, which classifies the most severe injury in each collision incident into five ordered categories. These categories progress from "None" (no significant injury), through "Minimal" (minor scrapes or bruises requiring no medical attention), "Minor" (requiring some medical attention but no hospitalization), "Major" (requiring hospitalization), to "Fatal" (death occurring within 30 days of the collision).

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-outcome
#| fig-cap: "Outcome frequency"
cleaned_data %>%
  ggplot(aes(x = injury, fill = injury)) +
  geom_bar() +
  scale_fill_brewer(palette = "RdYlBu") +
  labs(x = "Injury Severity",
       y = "Number of Collisions") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = scales::comma)
```
@fig-outcome The primary outcome variable in our analysis is collision severity, represented through an ordered factor ranging from "None" to "Fatal". It illustrates the distribution of injury severity across all recorded incidents. The data reveals an interesting bimodal distribution in collision outcomes, with concentrations at both ends of the severity spectrum. Major injuries and no injuries represent the most frequent outcomes, accounting for 40.1% and 38.9% of incidents respectively. Fatal collisions comprise 6.3% of the dataset, while minor and minimal injuries together represent approximately 15% of cases.

## Predictor variables {#sec-data-predictor}
This study incorporates thirteen predictor variables spanning temporal, environmental, infrastructural, and human factors. Temporal aspects are captured through date and time of collision occurrence. Environmental conditions are measured through three variables: visibility (recording conditions such as Clear, Rain, or Snow), light (indicating natural light conditions like Daylight, Dusk, or Dark), and rdsfcond (road surface condition, noting whether the road was Dry, Wet, or covered in Snow/Ice). Infrastructure characteristics are represented by road_class (categorizing roads as Major Arterial, Minor Arterial, Collector, Local, or Expressway), accloc (specifying whether the collision occurred at an intersection, mid-block, or other location), and traffctl (indicating the type of traffic control present, such as signals or signs). The human element is captured through five variables: invage (age of persons involved), speeding (indicating whether speeding was a factor), ag_driv (noting aggressive driving behavior), alcohol (indicating alcohol involvement), and disability (noting any relevant disabilities). Finally, impactype describes the nature of the collision, such as rear-end, angle, or turning movement.

### Environmental Conditions
```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-visibility
#| fig-cap: "Proportion of collisions severity in different visibility conditions"
cleaned_data %>%
  ggplot(aes(x = visibility, fill = injury)) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "RdYlBu") +
  labs(x = "Visibility Condition",
       y = "Proportion of Collisions",
       fill = "Injury Severity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@fig-visibility Visibility factor plays a crucial role in collision outcomes. Figure 2 presents the relationship between visibility conditions and collision severity. Clear conditions account for the majority of recorded incidents (86.4%), while poor visibility conditions (including rain, snow, and freezing rain) represent 10.5% of cases. Notably, the proportion of severe injuries increases under poor visibility conditions, suggesting a significant relationship between environmental conditions and collision outcomes.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-road
#| fig-cap: "Frequency of different injury severity in different road class"
cleaned_data %>%
  ggplot(aes(x = road_class, fill = injury)) +
  geom_bar(position = "stack") +
  scale_fill_brewer(palette = "RdYlBu") +
  labs(x = "Road Classification",
       y = "Number of Collisions",
       fill = "Injury Severity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = scales::comma)
```

@fig-road Road classification and traffic control measures represent important infrastructure factors in our analysis. Figure 3 illustrates the distribution of collisions across different road types and their associated severity levels. Major arterial roads account for 71.3% of all collisions, followed by minor arterial roads at 15.7%. This distribution likely reflects both traffic volume and speed limits on these road types. The relationship between road classification and injury severity reveals important patterns for urban safety planning.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-visibilityandlight
#| fig-cap: "Frequency of different injury severity in different combination of visibility and light"
# Prepare data: Summarize counts and calculate proportions
interaction_data <- cleaned_data %>%
  group_by(visibility, light, injury) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(visibility, light) %>%
  mutate(proportion = count / sum(count)) %>%
  ungroup()

# Create a facet grid visualization with RdYlBu color palette
ggplot(interaction_data, aes(x = light, y = proportion, fill = injury)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ visibility, scales = "free_y") +
  scale_fill_brewer(palette = "RdYlBu") +
  labs(
    title = "Interaction Between Visibility and Light on Injury Severity",
    x = "Light Conditions",
    y = "Proportion of Collisions",
    fill = "Injury Severity"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@fig-visibilityandlight The facet grid visualization reveals the interaction between visibility and light conditions on injury severity. Under poor visibility (fog), dark conditions show a noticeably higher proportion of severe injuries (Fatal, Major) compared to daylight, indicating the compounded risk of low visibility and inadequate lighting. In contrast, clear visibility with daylight consistently shows the lowest proportions of severe injuries, emphasizing the protective role of good environmental conditions. These findings will help guide the inclusion of visibility and light in the predictive model, ensuring it accounts for the impact of environmental factors on injury outcomes. This approach will improve the model's ability to identify scenarios with higher risks of severe injuries.

### Driver conditions
```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-behaviors
#| fig-cap: "Proportion of Injury Severity by whether Drivers have bad Behaviors"
# Prepare data: Summarize proportions for behaviors and injury severity
proportional_data <- cleaned_data %>%
  pivot_longer(cols = c(speeding, alcohol, ag_driv), names_to = "behavior", values_to = "status") %>%
  group_by(behavior, status, injury) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(behavior, status) %>%
  mutate(proportion = count / sum(count)) %>%
  ungroup()

# Create line chart for proportions
ggplot(proportional_data, aes(x = injury, y = proportion, group = status, color = status)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  facet_wrap(~ behavior, scales = "free_y") +
  labs(
    title = "Proportion of Injury Severity by Driver Behaviors (Yes/No)",
    x = "Injury Severity",
    y = "Proportion of Collisions",
    color = "Behavior Status"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )
```
@fig-behaviors The proportional line chart compares injury severity distributions for drivers involved in risky behaviors (speeding, alcohol use, aggressive driving) versus those who were not. It reveals that drivers engaged in these behaviors often have higher proportions of severe injuries, such as Fatal or Major, compared to those who did not. However, the results may be influenced by sample size imbalances, as there are likely fewer drivers in the "Yes" group for alcohol or aggressive driving, which could exaggerate proportions and introduce bias. Despite these limitations, the chart provides valuable insights into the heightened risks associated with these behaviors

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-impacttype
#| fig-cap: "Injury proportion in different impactypes"
# Group and calculate proportions
proportional_data <- cleaned_data %>%
  group_by(impactype, injury) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(impactype) %>%
  mutate(proportion = count / sum(count)) %>%
  ungroup()

# Filter top 5 impact types based on total count for clarity
top_impact_types <- proportional_data %>%
  group_by(impactype) %>%
  summarise(total_count = sum(count)) %>%
  arrange(desc(total_count)) %>%
  slice(1:5) %>%
  pull(impactype)

filtered_data <- proportional_data %>%
  filter(impactype %in% top_impact_types)

# Create the proportion-based bar chart with discrete color palette
ggplot(filtered_data, aes(x = impactype, y = proportion, fill = injury)) +
  geom_bar(stat = "identity", position = position_stack()) +
  scale_fill_brewer(palette = "RdYlBu") +
  labs(
    title = "Proportion of Injury Severity by Impact Type",
    x = "Impact Type",
    y = "Proportion of Collisions",
    fill = "Injury Severity"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@fig-impacttype The stacked bar chart reveals the proportion of injury severities across the top five collision impact types, providing insight into how different types of collisions relate to injury outcomes. For instance, pedestrian collisions show a significantly higher proportion of severe injuries (Fatal and Major), while rear-end collisions are dominated by minor or no injuries. This exploration is critical for our paper as it identifies which collision types pose the greatest risk of severe outcomes, helping us understand the underlying patterns of injury severity. By focusing on this relationship, we address a core question of our study: how specific predictors, such as impact type, influence injury severity.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-violin
#| fig-cap: "Age Distribution by Injury Severity"
# Create the violin plot
ggplot(cleaned_data, aes(x = injury, y = invage)) +
  geom_violin(scale = "count", trim = FALSE, fill = "lightblue") +
  stat_summary(fun = median, geom = "point", size = 2, color = "red") +
  labs(
    title = "Age Distribution by Injury Severity",
    x = "Injury Severity",
    y = "Age"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
@fig-violin The violin plot may show younger individuals having a higher density in some severer injury category such as "Major", indicating they are overrepresented in severe collisions. This trend can result from several factors: younger people are often more active on the roads, increasing their exposure to risk, and they may engage in riskier behaviors like speeding or distracted driving. Their lack of driving experience and the use of older, less safe vehicles can also contribute to more severe outcomes.

\newpage

# Model {#sec-model}
This study employs a two-stage modeling approach to understand and predict traffic injury outcomes, a model for injury occurrence prediction and another model for injury severity prediction when injuries occur. Both models were implemented using the brms package [@brms] in R, leveraging its robust framework for Bayesian regression modeling. The dual-model Bayesian framework precisely analyze traffic injuries, separating the prediction of injury occurrence from injury severity assessment. This approach allows us to isolate and examine the distinct factors that influence whether an injury occurs and, subsequently, how severe that injury becomes. By developing separate models for these two aspects, we can better understand the complex dynamics of traffic accidents and provide more targeted insights for safety interventions.

## Model set-up {#sec-model-setup}
### Injury occurence model
For our injury occurrence model, we implemented a fundamental binary transformation that distinguishes between collisions resulting in any injury and those without injury. In this classification scheme, we designated all collisions with "None" injury classification as our negative outcome ("No injury occurred"). Conversely, we grouped all other injury levels - Minimal, Minor, Major, and Fatal - into our positive outcome ("Injury occurred"). This transformation collapses the rich variety of injury outcomes into a simple binary indicator focusing solely on injury occurrence.

This model addresses the binary outcome of injury occurrence through a Bayesian logistic regression framework. Let $Y_i$ represent the injury outcome for observation $i$, where $Y_i = 1$ indicates an injury occurred and $Y_i = 0$ indicates no injury. The probability of injury is modeled through a Bernoulli distribution:

$$Y_i = 
\begin{cases} 
0 & \text{if no injury occurred (original classification: None)} \\ 
1 & \text{if any injury occurred (original classifications: Minimal, Minor, Major, or Fatal)}
\end{cases}$$

**\begin{align}Y_i \sim \text{Bernoulli}(p_i)\end{align}**

**$\text{logit}(p_i) = \beta_0 + \beta_1\text{Visibility}_i + \beta_2\text{Light}_i + \beta_3\text{RoadCondition}_i + \beta_4\text{RoadClass}_i + \beta_5\text{Location}_i + \beta_6\text{TrafficControl}_i + \beta_7\text{RushHour}_i + \beta_8\text{Weekend}_i + \beta_9\text{ImpactType}_i$**

The occurrence model prioritizes environmental and infrastructural factors, drawing on established traffic safety literature and the patterns observed in our exploratory data analysis. Environmental conditions are represented through visibility, light conditions, and road surface state, each categorized to capture fundamental differences while maintaining model parsimony. We consolidate visibility into a binary indicator (Good/Poor) based on detailed weather conditions, simplify lighting into day/night categories from six original conditions, and classify road conditions as good or poor based on surface characteristics.
Infrastructure factors incorporate the physical and operational aspects of the road environment. Road classification follows a hierarchical structure (Major/Minor/Local) that reflects traffic volume and design standards. Location type captures the crucial distinction between intersections, private drives, and other locations, while traffic control variables account for the presence and type of traffic management systems. Additionally, temporal factors such as rush hour periods and weekend indicators are included to capture systematic variations in traffic patterns.

### Injury severity model
Our severity model employs a more nuanced approach, focusing exclusively on collisions where injuries occurred. We first excluded all cases classified as "None" injury, as these fall outside the scope of severity analysis. Among injury cases, we created a binary classification that distinguishes between severe and minor injuries. We classified "Fatal" and "Major" injuries as "Severe" outcomes, while grouping "Minimal" and "Minor" injuries as "Minor" outcomes.

For cases where injuries occur, our second model examines severity levels using an ordered logistic regression approach. The severity outcome $S_i$ is modeled as:

$$S_i = 
\begin{cases}
0 & \text{if injury was less severe (original classifications: Minimal, Minor)} \\ 
1 & \text{if injury was severe (original classifications: Major, Fatal)}
\end{cases}$$

**\begin{align}S_i \sim \text{Bernoulli}(q_i)\end{align}**

**$\text{logit}(q_i) = \alpha_0 + \alpha_1\text{Speeding}_i + \alpha_2\text{AggressiveDriving}_i + \alpha_3\text{Alcohol}_i + \alpha_4\text{Disability}_i + \alpha_5\text{AgeGroup}_i + \alpha_6\text{Impact}_i + \alpha_7\text{Light}_i + \alpha_8\text{RoadClass}_i$**

The severity model emphasizes driver behavioral and impact factors that influence injury outcomes once an accident occurs. Driver behavior variables include binary indicators for speeding, aggressive driving, alcohol involvement, and driver disability. Age groups are categorized to capture the varying vulnerability and risk patterns across different life stages, while maintaining sufficient group sizes for reliable estimation.
Impact characteristics play a crucial role in determining injury severity. The model incorporates impact type as a categorical variable distinguishing between pedestrian, cyclist, rear-end, and angle collisions, each representing distinct injury risk patterns. Environmental conditions are represented through light conditions and road classification, which can moderate the relationship between impact forces and resulting injuries.

## Model validation {#sec-model-validate}
Our validation approach employs a sophisticated stratified sampling methodology, dividing the data into training (70%) and testing (30%) sets while carefully maintaining the original class proportions. This strategy is particularly crucial given the inherently imbalanced nature of traffic injury data. The validation framework encompasses multiple complementary approaches, including confusion matrix analysis for understanding classification performance, receiver operating characteristic (ROC) curve analysis for assessing discriminative ability. We further ensure model stability through comprehensive cross validation procedures. Prior to model fitting, we conducted thorough multicollinearity assessments using Variance Inflation Factors, confirming the independence of our predictor variables.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-posterior
#| fig-cap: "Posterior Predictive Check for Injury Occurrence Model"
pp_check(occurrence_model, type = "bars", ndraws = 100) +
  ggtitle("Posterior Predictive Check for Injury Occurrence Model")
```
@fig-posterior The posterior predictive check plot demonstrates the model's ability to replicate the binary nature of injury occurrence data. The blue dots (yrep) represent simulated data from the posterior predictive distribution, while the light blue bars (y) show the actual observed data distribution. The plot reveals two clear peaks around 0 and 1, corresponding to no-injury and injury cases respectively, with the simulated data closely matching the observed distribution. The alignment between predicted and actual counts at both peaks suggests that the model effectively captures the underlying binary structure of injury occurrences while maintaining appropriate uncertainty in its predictions.

\newpage

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-trace_o1
#| fig-cap: "Trace Plots for first 7 parameters in Injury Occurrence Model"
# First 7 parameters
mcmc_trace(as_draws(occurrence_model), 
          regex_pars = "b_(Intercept|visibilityPoor|lightNight|rdsfcondPoor|road_classMajor|road_classMinor|acclocOther)") +
          ggtitle("Trace Plots for Injury Occurrence Model - Part 1")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-trace_o2
#| fig-cap: "Trace Plots for middle 7 parameters in Injury Occurrence Model"
# Next 7 parameters
mcmc_trace(as_draws(occurrence_model), 
          regex_pars = "b_(acclocPrivate|traffctlOther|traffctlSignal|traffctlStop|impactypeApproaching|impactypeCyclistCollisions|impactypeOther)") +
          ggtitle("Trace Plots for Injury Occurrence Model - Part 2")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-trace_o3
#| fig-cap: "Trace Plots for last 6 parameters in Injury Occurrence Model"
# Last 8 parameters
mcmc_trace(as_draws(occurrence_model), 
          regex_pars = "b_(impactypePedestrianCollisions|impactypeRearEnd|impactypeSideswipe|impactypeSMVOther|impactypeSMVUnattendedVehicle|impactypeTurningMovement)") +
          ggtitle("Trace Plots for Injury Occurrence Model - Part 3")
```

@fig-trace_o1, @fig-trace_o2, @fig-trace_o3 The trace plots display stable mixing patterns across all model parameters over iterations. The four MCMC chains, shown in different blue shades, consistently overlap and explore similar value ranges, indicating good convergence. The stable patterns and lack of trending in the sampling trajectories suggest reliable parameter estimation. Environmental factors show relatively narrow ranges, while impact type coefficients exhibit wider ranges, reflecting their relative importance in predicting injury occurrence.

\newpage

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-trace_s1
#| fig-cap: "Trace Plots for first six parameters in Injury Secerity Model"
# First plot - first group of parameters
mcmc_trace(as_draws(severity_model), 
          regex_pars = "b_(Intercept|speeding|ag_driv|alcohol|disability|age)") +
          ggtitle("Trace Plots for Injury Severity Model - Part 1")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-trace_s2
#| fig-cap: "Trace Plots for rest parameters in Injury Secerity Model"
# Second plot - second group of parameters
mcmc_trace(as_draws(severity_model), 
          regex_pars = "b_(impact|light|road)") +
          ggtitle("Trace Plots for Injury Severity Model - Part 2")
```
@fig-trace_s1, @fig-trace_s2 The trace plots for various model parameters display stable MCMC sampling convergence across all four chains (represented in different shades of blue). Each parameter's trace shows consistent exploration within its respective range without any concerning patterns or trends, indicating good mixing of the chains. The overlapping nature of the chains and their stationary behavior throughout the sampling period provides strong evidence that the model has reached convergence and the posterior distributions are reliable estimates of the true parameter values.

\newpage

## Model assumption and limitation {#sec-model-assumpandlimit}
Both models operate under several key assumptions that warrant discussion. The independence of observations assumption may be partially violated due to spatial and temporal clustering of accidents, though our inclusion of location and time-based variables helps mitigate this concern. While multicollinearity was assessed and addressed during variable selection, some residual correlation between predictors may remain.

A notable limitation is the potential for unmeasured confounding factors, particularly in the severity model where detailed information about vehicle characteristics and exact impact speeds is often unavailable. The models also assume static relationships between predictors and outcomes over time, which may not hold in the presence of evolving vehicle safety technology and road design practices. Additionally, the reliance on reported data introduces potential selection bias, as minor accidents may be underreported.

## Alternative model consideration {#sec-model-alternate}
During the model development process, we evaluated several alternative specifications before settling on the current approach. A multinomial severity model was initially considered to capture finer gradations of injury severity but was ultimately rejected due to class imbalance issues and the practical utility of a binary severity classification. We explored mixed-effects models incorporating random effects for locations but found insufficient location-specific data to justify this complexity.

The final dual-model approach was selected for its balance of theoretical grounding, practical utility, and statistical robustness. By separating occurrence and severity, we maintain model interpretability while capturing the distinct processes that govern each outcome.

\newpage

# Results {#sec-result}
## Model result
Our dual-model analysis of traffic injuries reveals distinct patterns in both injury occurrence and severity prediction, with both models demonstrating strong predictive performance across different scenarios and conditions. Key findings from our models reveal the relative importance of different predictors in determining both injury occurrence and severity, and we summarise it in the follow coefficients tables and ROC curve plot.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: tbl-occurrence
#| tbl-cap: "Occurrence Model Coefficients"
# For occurrence model
occurrence_summary <- as.data.frame(summary(occurrence_model)$fixed)
occurrence_table <- occurrence_summary %>%
  rownames_to_column("Variable") %>%
  select(Variable, Estimate, `Est.Error`, `l-95% CI`, `u-95% CI`, `Bulk_ESS`, `Tail_ESS`) %>%
  rename(
    'Std. Error' = 'Est.Error',
    'Lower CI' = 'l-95% CI',
    'Upper CI' = 'u-95% CI',
    'Bulk ESS' = 'Bulk_ESS',
    'Tail ESS' = 'Tail_ESS'
  )

# Create formatted tables
kable(occurrence_table,
      format = "pipe",
      digits = 3)
```

@tbl-occurrence The Occurrence Model uncovers several noteworthy patterns in predicting the likelihood of injuries in traffic collisions. Among the environmental factors, nighttime conditions (0.131) are associated with a higher likelihood of injury, highlighting the risks associated with reduced visibility and nighttime driving.
Impact types emerge as the most influential predictors. Single motor vehicle (SMV) incidents show the strongest positive association with injury occurrence (1.224), likely reflecting the dynamics of these collisions, such as loss of control or impacts with stationary objects. Interestingly, cyclist (-0.773) and pedestrian (-0.772) collisions exhibit substantial negative associations with injury occurrence. This might seem contradictory to the findings in our Severity Model, where these types of collisions are strongly linked to severe injuries. However, the explanation may lie in the polarizing nature of these incidents many result in no injury at all or extreme injuries, which could skew their average likelihood of injury occurrence.
Most infrastructure-related variables, such as road classification and traffic control measures, show relatively modest or non-significant effects, with confidence intervals crossing zero. This suggests that their influence on injury occurrence might depend heavily on specific contexts or interactions with other factors.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: tbl-severity
#| tbl-cap: "Severity Model Coefficients"
# For severity model
severity_summary <- as.data.frame(summary(severity_model)$fixed)
severity_table <- severity_summary %>%
  rownames_to_column("Variable") %>%
  select(Variable, Estimate, `Est.Error`, `l-95% CI`, `u-95% CI`, `Bulk_ESS`, `Tail_ESS`) %>%
  rename(
    'Std. Error' = 'Est.Error',
    'Lower CI' = 'l-95% CI',
    'Upper CI' = 'u-95% CI',
    'Bulk ESS' = 'Bulk_ESS',
    'Tail ESS' = 'Tail_ESS'
  )

cat("\nTable 5: Coefficients for Severity Model\n")
kable(severity_table,
      format = "pipe",
      digits = 3)
```

@tbl-severity The results of the Severity Model reveal important insights into the factors contributing to severe injuries in motor vehicle collisions. Collisions involving cyclists (2.31) and pedestrians (1.73) stand out as the most significant predictors of severe outcomes, emphasizing the heightened vulnerability of these road users. Senior individuals are also more likely to experience severe injuries, reflecting the increased physical fragility of this age group. Interestingly, collisions on minor roads are associated with a higher likelihood of severe injuries, possibly due to less stringent traffic control measures or more complex road layouts. On the other hand, factors like speeding and alcohol involvement, while often associated with risky driving behaviors, did not show a strong connection to injury severity in this dataset.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-roc
#| fig-cap: "ROC Curve for Injury Severity Model"
# Predicted Probabilities
severity_probs <- fitted(severity_model, newdata = severity_test, summary = FALSE)

# Average over posterior samples
severity_mean_probs <- colMeans(severity_probs)

# ROC Curve
roc_severity <- roc(severity_test$injury_severity, severity_mean_probs)
plot(roc_severity, main = "ROC Curve for Injury Severity Model")
```

@fig-roc The ROC curve demonstrates the model's discriminative ability across different classification thresholds. The curve rises sharply from the origin and maintains considerable distance from the diagonal reference line (which represents random guess), indicating strong predictive power. The smooth progression of the curve suggests stable performance across various sensitivity-specificity trade-offs, while its position well above the diagonal line confirms the model's robust ability to distinguish between injury severity classes.

## Prediction example
Our prediction analysis utilizes scenario-based datasets (occurrence_scenarios and severity_scenarios) constructed with expand_grid to systematically explore various combinations of environmental, infrastructural, and behavioral conditions. By inputting these scenarios into our models, we generate predicted probabilities of injury occurrence and severity across a comprehensive range of real-world situations.

To evaluate the injury severity model's predictive performance, we compared actual outcomes with predicted probabilities using the test dataset. The follow plot displays the relationship between actual injury severity outcomes and the predicted probabilities generated by the model.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-avp
#| fig-cap: "Actual vs. Predicted Probabilities (Injury Severity Model)"
# Predicted Probabilities on Test Data
severity_test$predicted_prob <- fitted(
  severity_model,
  newdata = severity_test,
  summary = TRUE
)[, "Estimate"]

# Convert Actual Outcomes to Numeric
severity_test$actual <- ifelse(severity_test$injury_severity == "Severe", 1, 0)

# Scatter Plot
ggplot(severity_test, aes(x = predicted_prob, y = actual)) +
  geom_jitter(height = 0.05, alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(
    title = "Actual vs. Predicted Probabilities (Injury Severity Model)",
    x = "Predicted Probability",
    y = "Actual Outcome"
  ) +
  theme_minimal()
```

@fig-avp The actual versus predicted probabilities plot for the injury severity model reveals a strong predictive performance, with observations clustering predominantly at either high (around 1.0) or low (around 0.0) probabilities. The blue diagonal line represents perfect prediction, and the pattern of actual outcomes shows clear separation between severity classes, indicating the model's ability to effectively discriminate between minor and severe injuries. This binary clustering pattern is characteristic of a well-calibrated binary classification model.

The plot demonstrates that higher predicted probabilities correspond to actual severe injuries, indicating the model's effectiveness in predicting injury severity. This alignment between predicted probabilities and actual outcomes confirms the model's practical utility in identifying high-risk scenarios.
Our prediction results are stored in the "models" directory as occurrence_predictions.parquet and severity_predictions.parquet.

# Discussion {#sec-discussion}
## Data quality and limitations
Our analysis faces several important limitations related to data quality and collection. The primary collision dataset from opendatatoronto [@opendatatoronto] contains inherent biases in injury reporting. According to the National Highway Traffic Safety Administration's recent report [@nhtsa2023], up to 40% of minor injury accidents go unreported, particularly in cases involving single vehicles or where parties choose to handle incidents privately. This underreporting likely creates a selection bias towards more severe accidents in our dataset, potentially affecting our model's predictions for minor injury scenarios.

Additionally, the geographic distribution of our data shows significant imbalance. Urban areas are typically overrepresented due to better reporting infrastructure and higher population density, while rural accidents may be underreported or documented with less detail. The Insurance Institute for Highway Safety [@iihs2023] notes that rural accidents, despite being fewer in number, often result in more severe injuries due to factors such as emergency response times and road conditions - aspects our current data structure may not fully capture.

## Model tradeoffs
Our choice of a Bayesian framework, while powerful for uncertainty quantification, introduces its own set of tradeoffs. The computational intensity of MCMC sampling, particularly with our large dataset and multiple chains (4 chains with 4,000 iterations each), requires significant processing time. As discussed in recent methodological work by Gelman [@gelman2023], the balance between model complexity and computational feasibility often requires compromises in the number of parameters or interactions that can be practically included.

The separation of our analysis into occurrence and severity models, while theoretically sound, may miss important interactions between these aspects. Several factors influencing occurrence and severity are often interconnected in ways that separate models might not fully capture. For instance, our models treat environmental conditions like visibility as independent factors, when in reality they may have compound effects that our current approach cannot quantify.

The model structure also assumes temporal independence between accidents, an assumption that likely doesn't reflect reality. Accident patterns often show temporal correlation, particularly during adverse weather conditions or holiday periods. Our current framework cannot adequately capture these temporal dependencies, potentially overlooking important patterns in accident occurrence and severity.

## Variable handling challenges
A significant limitation lies in our variable categorization approach. Environmental conditions are simplified into binary categories (e.g., "Good"/"Poor" visibility, "Day"/"Night" lighting) to maintain model stability and interpretability. However, this simplification may mask important nuances in intermediate conditions. For example, the effect of partial visibility or dusk conditions might differ significantly from either clear or poor visibility, but our current categorization cannot capture these distinctions.

The handling of behavioral factors in our models also presents challenges. While we account for factors like speeding and aggressive driving, the binary nature of these variables may oversimplify complex behavioral patterns. Additionally, our reliance on reported data means we can only capture behaviors that were documented at the time of the accident, potentially missing important precursor behaviors or conditions.

## Future research directions/improvements
Several promising directions for improvement emerge from our analysis. First, the integration of continuous monitoring data could provide more granular insights into accident patterns. This could include real-time traffic flow data, weather conditions, and road surface measurements, allowing for more nuanced understanding of how environmental conditions affect accident risk.

Alternative modeling approaches, such as hierarchical spatial models or dynamic time series analyses, could better capture the complex dependencies in accident data. The incorporation of spatial correlation structures could improve our understanding of location-specific risks as current research suggests [@plosone_spatial_models], while time-series components could better account for temporal patterns in accident occurrence.

The development of more sophisticated variable interactions, particularly between environmental and behavioral factors, represents another important avenue for future research. Our current models treat these factors largely independently, but evidence from the national academies of sciences [@nationalacademies_ontologies] suggests complex interactions between driver behavior, road conditions, and accident outcomes that warrant more detailed examination.

To address the identified limitations, future data collection efforts should prioritize standardized reporting protocols and the integration of automated data collection systems. This could help reduce reporting biases and provide more complete coverage of accident scenarios.

\newpage

\appendix

# Appendix {-}
## Data collecting methodology {#sec-data-methodology}
The City of Toronto's Motor Vehicle Collision database involving Killed or Seriously Injured Persons (KSI) is collected through a systematic process involving the Toronto Police Service (TPS) as the primary data collector [@toronto_collision_data]. When a collision occurs involving killed or seriously injured persons, police officers must complete a standardized Motor Vehicle Accident Report form. This data is then processed and maintained by the City of Toronto's Transportation Services Division. The collection process follows a standardized protocol outlined in the Motor Vehicle Collision Report (MVCR) manual [@mvcr_guide], which ensures consistency in reporting and classification across all collision events.

The data collection process begins when police officers attend a collision scene and complete the Motor Vehicle Collision Report form (SR-LD-401), which follows a standardized protocol detailed in the Motor Vehicle Collision Report Manual [@mvcr_guide]. This systematic approach ensures consistency in data collection across different incidents and officers. However, this methodology introduces an important selection bias - the dataset only includes collisions where police officers were present and generated a report. Consequently, incidents where involved parties did not contact police or left the scene before police arrival are systematically excluded from the dataset.

A critical aspect of the data collection methodology is its focus on Killed or Seriously Injured (KSI) events. The dataset defines major injury as "a non-fatal injury severe enough to require the injured person to be admitted to hospital, even if only for observation." This includes fractures, internal injuries, severe cuts, crushing, burns, concussion, and severe general shocks. Fatal injuries are defined as deaths occurring within 366 days as a result of the collision, explicitly excluding deaths from natural causes or suicide [@toronto_police_services_ksi_guide]. This definition creates a clear boundary for data inclusion but may underestimate the total impact of traffic collisions by excluding less severe injuries that still have significant societal costs.

The dataset's structure reveals a careful attention to detail in recording various aspects of each collision. Each record represents a person involved in a KSI collision event, regardless of their injury level. This means that a single collision event may have multiple records, one for each person involved. While this provides rich detail about all participants in serious collisions, it creates challenges for analyzing collision events as distinct units without careful data processing to account for duplicate event numbers (ACCNUM).

A significant methodological strength is the comprehensive categorization system used for various collision attributes. The dataset includes detailed classifications for environmental conditions, road surface conditions, collision types, and participant behaviors [@toronto_police_services_ksi_guide]. However, this strength is partially offset by the subjective nature of some classifications. For instance, the assessment of "apparent driver action" or "driver condition" relies heavily on officer judgment at the scene, introducing potential observer bias.

Another notable aspect of the methodology is its handling of spatial data. Each collision is geocoded with latitude and longitude coordinates, enabling spatial analysis. However, the documentation acknowledges that some location data may be approximate, particularly for collisions occurring between major intersections or in areas without clear address points.
The reason is the balance between public data transparency and privacy protection in administrative data collection. The feature of this methodology is the deliberate offsetting of collision locations to the nearest road intersection node, in order to safeguard the privacy of involved parties [@toronto_collision_data]. This geographic anonymization technique, while serving its primary purpose of privacy protection, introduces spatial imprecision that affects the dataset's geographical analysis capabilities. Consequently, aggregated statistics at the Division and Neighbourhood levels may not precisely reflect the true spatial distribution of collisions. The Toronto Police Service explicitly acknowledges these limitations, cautioning users about potential discrepancies in accuracy, completeness, and timeliness when compared to other crime data sources.

The dataset's temporal coverage and update frequency also merit consideration. While historical data is preserved, there is typically a lag time between incident occurrence and data availability in the public dataset, primarily due to the need for verification and processing of police reports. This delay affects the dataset's utility for real-time analysis but ensures higher data quality through proper verification procedures.

## Idealized methodology {#sec-data-idealized}
An idealized methodology for motor vehicle collision data collection must fundamentally reimagine how we gather, validate, and integrate traffic incident information while maintaining robust privacy protections. Drawing from advancements in digital technology and data science, I propose a comprehensive modernization of the current system.
The cornerstone of this enhanced methodology would be transitioning from paper-based documentation to a sophisticated digital data collection framework. Police officers would utilize specialized mobile devices equipped with purpose-built software that guides them through systematic data collection at collision scenes [@digital_transformation]. This digital transformation would significantly reduce transcription errors while enabling real-time data validation - a crucial improvement over the current SR-LD-401 form system.

Privacy protection in location data requires particular attention and refinement. The current practice of offsetting collision locations to the nearest intersection, while serving its primary purpose, introduces unnecessary spatial imprecision. Advanced geographic anonymization techniques could better balance privacy concerns with analytical precision. For instance, implementing differential privacy algorithms could add calibrated noise to geographic coordinates while preserving statistical accuracy for analysis purposes [@geographic_anonymization].

Environmental context represents another area where current methodologies rely heavily on subjective officer assessment. A modernized system would automatically integrate data from multiple validated sources, creating a more objective and comprehensive picture of collision conditions. This would include real-time weather data from official weather stations, traffic volume information from road sensors, and current road condition data from municipal databases. Such integration would provide a more reliable environmental context for each incident.

Documentation of physical evidence and collision dynamics would benefit from technological advancement as well. Standardized 360-degree photography and LiDAR scanning could provide precise measurements and objective documentation of vehicle damage and scene characteristics [@collision_reconstruction]. Computer vision algorithms could process this visual data to create standardized collision reconstructions, reducing reliance on subjective interpretation and improving consistency across incidents.

To address the current limitation of potential underreporting, the system should incorporate a multi source verification framework. By cross referencing police reports with emergency medical services records, hospital admission data, and insurance claims, while employing privacy-preserving record linkage techniques and the system would capture a more complete picture of collision incidents and their outcomes.
This idealized methodology may require significant investment in technology infrastructure and training, but it represents a necessary evolution in traffic collision data collection, which resulting improvements in data quality and data completeness.

## Model details {#sec-model-details}

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-ppo
#| fig-cap: "Examining how the Injury Occurrence Model fits)"
pp_check(occurrence_model, nsamples = 100) +
  ggtitle("Posterior Predictive Check for Injury Occurrence Model")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-eeo
#| fig-cap: "Summary for 95% confidence intervals of estimates in occurrence model"
modelplot(occurrence_model, coef_omit = "(Intercept)", conf_level = 0.95) +
  ggtitle("Estimated Effects on injury Occurrence") +
  theme_minimal()
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: tbl-occurrence_coef
#| tbl-cap: "Summary table for estimates of occurrence model"
# Extract the fixed effects summary
fixed_effects <- summary(occurrence_model)$fixed

# Display as a table with kable
kable(fixed_effects, 
      format = "markdown",  # Choose the desired format ("markdown", "html", or "latex")
      digits = 3,           # Round numbers to 3 decimal places
      col.names = c("Estimate", "Est. Error", "l-95% CI", "u-95% CI", "Rhat", "Bulk ESS", "Tail ESS"), # Rename columns
      caption = "Summary of Fixed Effects for the Model")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-pps
#| fig-cap: "Examining how the Injury Severity Model fits)"
pp_check(severity_model, nsamples = 100) +
  ggtitle("Posterior Predictive Check for Injury Occurrence Model")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-ees
#| fig-cap: "Summary for 95% confidence intervals of estimates in severity model"
modelplot(severity_model, coef_omit = "(Intercept)", conf_level = 0.95) + 
  ggtitle("Estimated Effects on injury severity") +
  theme_minimal()
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| label: tbl-severity_coef
#| tbl-cap: "summary table for estimates of severity model"
# Extract the fixed effects summary
fixed_effects <- summary(severity_model)$fixed

# Display as a table with kable
kable(fixed_effects, 
      format = "markdown",  # Choose the desired format ("markdown", "html", or "latex")
      digits = 3,           # Round numbers to 3 decimal places
      col.names = c("Estimate", "Est. Error", "l-95% CI", "u-95% CI", "Rhat", "Bulk ESS", "Tail ESS"), # Rename columns
      caption = "Summary of Fixed Effects for the Model")
```

\newpage

# References

### 2024-11-26 scripts/05-model_data.R
Update model construction script and store model related data

@@ -1,0 +1,361 @@
+#### Preamble ####
+# Purpose: Construct two bayesian models for the prediction of injury occurrence and severity
+# Author: Yitong Wang
+# Date: 21 November 2024
+# Contact: stevenn.wang@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: 
+#   - 03-clean_data.R must have been run
+# Any other information needed? Make sure you are in the `toronto_collision_analysis` rproj
+
+# Load packages
+library(tidyverse)
+library(brms)
+library(here)
+library(tidybayes)
+library(bayesplot)
+library(pROC)
+
+# Set Seed for Reproducibility
+set.seed(777)
+
+# Data Preparation
+# Prepare data for both models with different predictor sets
+
+# Model 1: Injury Occurrence (focuses on circumstantial and infrastructural factors)
+occurrence_data <- cleaned_data %>%
+  mutate(
+    # Binary outcome
+    injury_occurred = case_when(
+      injury == "None" ~ "No",
+      TRUE ~ "Yes"
+    ) %>% factor(levels = c("No", "Yes")),
+    
+    # Environmental conditions
+    visibility = case_when(
+      visibility == "Clear" ~ "Good",
+      TRUE ~ "Poor"
+    ) %>% factor(),
+    
+    light = case_when(
+      light %in% c("Daylight", "Daylight, artificial") ~ "Day",
+      TRUE ~ "Night"
+    ) %>% factor(),
+    
+    rdsfcond = case_when(
+      rdsfcond == "Dry" ~ "Good",
+      TRUE ~ "Poor"
+    ) %>% factor(),
+    
+    # Infrastructure factors
+    road_class = case_when(
+      road_class %in% c("Major Arterial", "Expressway") ~ "Major",
+      road_class %in% c("Minor Arterial", "Collector") ~ "Minor",
+      TRUE ~ "Local"
+    ) %>% factor(),
+    
+    # Location type
+    accloc = case_when(
+      str_detect(accloc, "Intersection") ~ "Intersection",
+      str_detect(accloc, "Private") ~ "Private",
+      TRUE ~ "Other"
+    ) %>% factor(),
+    
+    # Traffic control
+    traffctl = case_when(
+      traffctl == "Traffic Signal" ~ "Signal",
+      traffctl == "Stop Sign" ~ "Stop",
+      traffctl == "No Control" ~ "None",
+      TRUE ~ "Other"
[…]

### 2024-11-26 paper/paper.qmd
Refine paper content

@@
 author: Yitong Wang
-thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
+thanks: "Code and data are available at: [https://github.com/demainwang/Toronto_vehicle_collision_analysis](https://github.com/demainwang/Toronto_vehicle_collision_analysis)."
 date: today

### 2024-11-29 paper/paper.qmd
Refine the content of paper
Refine the paper and add some sections

--- 2024-11-26 paper/paper.qmd
@@
 ---
-title: "Environmental Conditions and Driver Behavior Shape Collision Outcomes in Toronto"
-subtitle: "A Two-Stage Bayesian Analysis Reveals Infrastructure Factors Predict Crash Occurrence While Driver Actions Determine Severity"
+title: "Environmental Conditions and Driver Behavior Affect Collision Outcomes in Toronto"
+subtitle: "A Two-Stage Bayesian Analysis Shows Infrastructure Factors Predict Crash Occurrence While Driver Actions Determine Severity"
 author: Yitong Wang
-thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
+thanks: "Code and data are available at: [https://github.com/demainwang/Toronto_vehicle_collision_analysis](https://github.com/demainwang/Toronto_vehicle_collision_analysis)."
 date: today
@@
 The first model predicts the probability of injury based on environmental and infrastructural factors, while the second estimates injury severity using driver characteristics and behavior.
-Analysis of historical collision data from 2006 to 2023 reveals that while environmental conditions primarily influence injury occurrence, driver behavior factors like speeding and alcohol use more strongly affect severity levels.
-Our dual model analysis providing evidence based guidance for targeting both infrastructure improvements and enforcement strategies to reduce severe collision outcomes."
+Analysis of historical collision data from 2006 to 2023 shows that while environmental conditions primarily influence injury occurrence, driver behavior factors like speeding and alcohol use more strongly affect severity levels.
+These findings demonstrate the need for a dual-pronged approach to urban safety through targeted infrastructure improvements in high-risk areas to prevent collisions, alongside enhanced enforcement strategies focusing on dangerous driving behaviors.
+"
 format: pdf
@@
 Vehicle collisions in urban environments represent a significant public health and safety challenge, with particular concern in large metropolitan areas like Toronto where complex traffic patterns and diverse environmental conditions create varying levels of risk.
-Understanding the factors that contribute to collision occurrence and severity is crucial for urban planning and public safety policy, yet traditional analyses often treat these as a single outcome rather than recognizing their distinct underlying processes.
-The increasing availability of detailed collision data, combined with advances in statistical modeling, presents an opportunity to develop more nuanced understandings of how different factors influence collision outcomes.
+Understanding the factors that contribute to collision occurrence and severity is important for urban planning and public safety policy, yet traditional analyses often treat these as a single outcome rather than recognizing their distinct underlying processes.
+The increasing availability of detailed collision data, combined with advances in statistical modeling, presents an opportunity to develop more precise understandings of how different factors influence collision outcomes.
 
@@
 
-Our analysis reveals several distinct and important patterns in how different factors influence collision outcomes.
+Our analysis shows several distinct and important patterns in how different factors influence collision outcomes.
 Environmental conditions emerged as the dominant predictors of whether an injury occurs in the first place - particularly reduced visibility during winter months and poor road surface conditions on major arterials.
-However, once an injury occurs, driver behavior becomes the crucial determinant of severity.
+However, once an injury occurs, driver behavior becomes a determinant of severity.
 Impaired driving and speeding showed strong associations with severe outcomes, especially during nighttime hours and weekends.
@@
 @sec-data presents the overview of data (@sec-data-overview), data cleaning process (@sec-data-cleaning), measurement (@sec-data-measure), also for explanations, descriptions, graph summaries of outcome (@sec-data-outcome) and predictor variables (@sec-data-predictor) of the study.
-@sec-model set up the model and explain the process detailed, including procedure of model set up (@sec-model-setup) and model validation (@sec-model-validate), and also mention the assumption and validation (@sec-model-assumpandlimit) of models.
+@sec-model set up the model and explain the process detailed, including procedure of model set up (@sec-model-setup) ,and also mention the assumption and validation (@sec-model-assumpandlimit) of models.
 In addition, (@sec-model-alternate) discuss the potential alternative models and its trade offs.
-Then, @sec-result reveal the prediction outcome and models' performance.
+Then, @sec-result presents the prediction outcome and models' performance.
 @sec-discussion discusses the relation between the data and real world, which talks about the data quality and bias, as well as the potential limitation of our models in a broader context.
 
-Appendix includes three parts: [Appendix -@sec-data-methodology] presents collision data methodology overview, [Appendix -@sec-data-idealized] provides detailed idealized methodology for collection improvement; [Appendix -@sec-model-details] presents additional summary and validation in modeling process.
+Appendix includes three parts: [Appendix -@sec-data-methodology] presents collision data methodology overview, [Appendix -@sec-data-idealized] provides detailed idealized methodology for collection improvement; [Appendix -@sec-model-validate_and_details] presents model validation (@sec-model-validate) and additional details in modeling process (@sec-model-detail).
 
@@
 In this study, we utilized statistical language R [@citeR] and several key packages for data processing, analysis, and visualization.
-The tidyverse package [@tidyverse] facilitated data manipulation, with dplyr [@dplyr] being particularly crucial for data transformation and filtering operations, while visualization was conducted using ggplot2 [@ggplot2].
+The tidyverse package [@tidyverse] facilitated data manipulation, with dplyr [@dplyr] for data transformation and filtering operations, while visualization was conducted using ggplot2 [@ggplot2].
 For efficient and consistent data import, we employed the readr package [@readr].
-For handling temporal aspects of collision data, we employes lubridate {@lubridate} to process dates and times.We used here [@here] for reproducible file path management, arrow [@arrow] for efficient data storage and retrieval, and knitr [@knitr] used to provide the better formatting output.
+For handling temporal aspects of collision data, we employes lubridate (@lubridate) to process dates and times.We used here [@here] for reproducible file path management, arrow [@arrow] for efficient data storage and retrieval, and knitr [@knitr] used to provide the better formatting output.
 For unit testing and code validation, we employed the testthat package [@testthat] which provided a robust framework for testing our functions and ensuring code reliability.
@@
 From the raw dataset, we selected 14 key variables most relevant to understanding collision severity patterns: date, time, road_class (road classification), accloc (accident location), traffctl (traffic control), visibility, light, rdsfcond (road surface condition), impactype (impact type), invage (involved person's age), injury (severity), speeding, ag_driv (aggressive driving), alcohol, and disability.
-Records with missing values in critical fields such as injury severity, visibility, or road classification were removed to ensure data quality.
+Records with missing values in important fields such as injury severity, visibility, or road classification were removed to ensure data quality.
 
@@
 This additional cleaning step ensures our analysis is based on collisions with well-documented characteristics, though it reduced our dataset by approximately 8% (1,057 records).
-This filtering helps maintain data quality and supports more reliable analysis of factor relationships with collision severity.
 
 ## Measurement {#sec-data-measure}
-The transformation of real-world collision incidents into structured data points follows a rigorous protocol established by Toronto Police Services.
+The transformation of real-world collision incidents into structured data follows a rigorous protocol established by Toronto Police Services.
 When a collision occurs, first responders and investigating officers collect data through a standardized Motor Vehicle Accident Report Form.
@@
 Some driver related factors present particular measurement challenges due to their complex nature.
-Alcohol involvement is measured through standardized tests when possible, and it providing more objective data.
+Alcohol involvement is measured through standardized tests if it is possible, and it providing more objective data.
 However, other driver conditions like fatigue or distraction often rely on post incident interviews and officer's observation, which introducing potential reporting biases.
 Additionally, factors like driver age and license status are verified through official documentation, ensuring accuracy in demographic data.
-The measurement process also captures infrastructure characteristics such as road classification and traffic control presence, which are determined through existing municipal records and on-site verification.
-This multi-faceted approach to data collection, while comprehensive, means that the quality and reliability of different variables may vary depending on the objectivity of their measurement methods.
+The measurement process also captures infrastructure characteristics such as road classification and traffic control presence, which are determined through existing municipal records and on site verification.
+Even this multi faceted approach makes data is well covered, it also means the quality and reliability of different variables may vary depending on the objectivity of their measurement methods.
 
@@
 It illustrates the distribution of injury severity across all recorded incidents.
-The data reveals an interesting bimodal distribution in collision outcomes, with concentrations at both ends of the severity spectrum.
+The data shows an interesting bimodal distribution in collision outcomes, with concentrations at both ends of the severity spectrum.
 Major injuries and no injuries represent the most frequent outcomes, accounting for 40.1% and 38.9% of incidents respectively.
@@
 
-@fig-visibility Visibility factor plays a crucial role in collision outcomes.
-Figure 2 presents the relationship between visibility conditions and collision severity.
+@fig-visibility Visibility factor plays a important role in collision outcomes.
+It presents the relationship between visibility conditions and collision severity.
 Clear conditions account for the majority of recorded incidents (86.4%), while poor visibility conditions (including rain, snow, and freezing rain) represent 10.5% of cases.
@@
 
-@fig-road Road classification and traffic control measures represent important infrastructure factors in our analysis.
-Figure 3 illustrates the distribution of collisions across different road types and their associated severity levels.
+@fig-road illustrates the distribution of collisions across different road types and their associated severity levels.
 Major arterial roads account for 71.3% of all collisions, followed by minor arterial roads at 15.7%.
 This distribution likely reflects both traffic volume and speed limits on these road types.
-The relationship between road classification and injury severity reveals important patterns for urban safety planning.
 
@@
   theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+  theme(axis.text.x = element_text(size = 6, angle = 45, hjust = 1))
 ```
 
-@fig-visibilityandlight The facet grid visualization reveals the interaction between visibility and light conditions on injury severity.
+@fig-visibilityandlight The facet grid visualization presents the interaction between visibility and light conditions on injury severity.
 Under poor visibility (fog), dark conditions show a noticeably higher proportion of severe injuries (Fatal, Major) compared to daylight, indicating the compounded risk of low visibility and inadequate lighting.
@@
 These findings will help guide the inclusion of visibility and light in the predictive model, ensuring it accounts for the impact of environmental factors on injury outcomes.
-This approach will improve the model's ability to identify scenarios with higher risks of severe injuries.
+
+\newpage
 
@@
 @fig-behaviors The proportional line chart compares injury severity distributions for drivers involved in risky behaviors (speeding, alcohol use, aggressive driving) versus those who were not.
-It reveals that drivers engaged in these behaviors often have higher proportions of severe injuries, such as Fatal or Major, compared to those who did not.
+It shows that drivers engaged in these behaviors often have higher proportions of severe injuries, such as Fatal or Major, compared to those who did not.
 However, the results may be influenced by sample size imbalances, as there are likely fewer drivers in the "Yes" group for alcohol or aggressive driving, which could exaggerate proportions and introduce bias.
-Despite these limitations, the chart provides valuable insights into the heightened risks associated with these behaviors
 
@@
 
-@fig-impacttype The stacked bar chart reveals the proportion of injury severities across the top five collision impact types, providing insight into how different types of collisions relate to injury outcomes.
+@fig-impacttype The stacked bar chart demonstrates the proportion of injury severities across the top five collision impact types, providing understand for how different types of collisions relate to injury outcomes.
 For instance, pedestrian collisions show a significantly higher proportion of severe injuries (Fatal and Major), while rear-end collisions are dominated by minor or no injuries.
-This exploration is critical for our paper as it identifies which collision types pose the greatest risk of severe outcomes, helping us understand the underlying patterns of injury severity.
-By focusing on this relationship, we address a core question of our study: how specific predictors, such as impact type, influence injury severity.
+
+\newpage
 
@@
 This study employs a two-stage modeling approach to understand and predict traffic injury outcomes, a model for injury occurrence prediction and another model for injury severity prediction when injuries occur.
-Both models were implemented using the brms package [@brms] in R, leveraging its robust framework for Bayesian regression modeling.
-The dual-model Bayesian framework precisely analyze traffic injuries, separating the prediction of injury occurrence from injury severity assessment.
+Both models were implemented using the brms package [@brms] in R, the dual-model Bayesian framework precisely analyze traffic injuries, separating the prediction of injury occurrence from injury severity assessment.
 This approach allows us to isolate and examine the distinct factors that influence whether an injury occurs and, subsequently, how severe that injury becomes.
-By developing separate models for these two aspects, we can better understand the complex dynamics of traffic accidents and provide more targeted insights for safety interventions.
+By developing separate models for these two aspects, we can better understand the complex dynamics of traffic accidents.
 
@@
 
-This model addresses the binary outcome of injury occurrence through a Bayesian logistic regression framework.
 Let $Y_i$ represent the injury outcome for observation $i$, where $Y_i = 1$ indicates an injury occurred and $Y_i = 0$ indicates no injury.
@@
 Road classification follows a hierarchical structure (Major/Minor/Local) that reflects traffic volume and design standards.
-Location type captures the crucial distinction between intersections, private drives, and other locations, while traffic control variables account for the presence and type of traffic management systems.
+Location type captures the distinction between intersections, private drives, and other locations, while traffic control variables account for the presence and type of traffic management systems.
 Additionally, temporal factors such as rush hour periods and weekend indicators are included to capture systematic variations in traffic patterns.
@@
 ### Injury severity model
-Our severity model employs a more nuanced approach, focusing exclusively on collisions where injuries occurred.
+Our severity model is focusing exclusively on collisions where injuries occurred.
 We first excluded all cases classified as "None" injury, as these fall outside the scope of severity analysis.
@@
 
-For cases where injuries occur, our second model examines severity levels using an ordered logistic regression approach. The severity outcome $S_i$ is modeled as:
+For cases where injuries occur, the severity outcome $S_i$ is modeled as:
 
@@
 Age groups are categorized to capture the varying vulnerability and risk patterns across different life stages, while maintaining sufficient group sizes for reliable estimation.
-Impact characteristics play a crucial role in determining injury severity.
+Impact characteristics play a imporatant role in determining injury severity.
 The model incorporates impact type as a categorical variable distinguishing between pedestrian, cyclist, rear-end, and angle collisions, each representing distinct injury risk patterns.
 Environmental conditions are represented through light conditions and road classification, which can moderate the relationship between impact forces and resulting injuries.
-
-## Model validation {#sec-model-validate}
-Our validation approach employs a sophisticated stratified sampling methodology, dividing the data into training (70%) and testing (30%) sets while carefully maintaining the original class proportions.
-This strategy is particularly crucial given the inherently imbalanced nature of traffic injury data.
-The validation framework encompasses multiple complementary approaches, including confusion matrix analysis for understanding classification performance, receiver operating characteristic (ROC) curve analysis for assessing discriminative ability.
-We further ensure model stability through comprehensive cross validation procedures.
-Prior to model fitting, we conducted thorough multicollinearity assessments using Variance Inflation Factors, confirming the independence of our predictor variables.
-
-```{r, echo=FALSE, message=FALSE, warning=FALSE}
-#| label: fig-posterior
-#| fig-cap: "Posterior Predictive Check for Injury Occurrence Model"
-pp_check(occurrence_model, type = "bars", ndraws = 100) +
-  ggtitle("Posterior Predictive Check for Injury Occurrence Model")
-```
-@fig-posterior The posterior predictive check plot demonstrates the model's ability to replicate the binary nature of injury occurrence data.
-The blue dots (yrep) represent simulated data from the posterior predictive distribution, while the light blue bars (y) show the actual observed data distribution.
-The plot reveals two clear peaks around 0 and 1, corresponding to no-injury and injury cases respectively, with the simulated data closely matching the observed distribution.
-The alignment between predicted and actual counts at both peaks suggests that the model effectively captures the underlying binary structure of injury occurrences while maintaining appropriate uncertainty in its predictions.
-
-\newpage
-
-```{r, echo=FALSE, message=FALSE, warning=FALSE}
-#| label: fig-trace_o1
-#| fig-cap: "Trace Plots for first 7 parameters in Injury Occurrence Model"
-# First 7 parameters
-mcmc_trace(as_draws(occurrence_model), 
-          regex_pars = "b_(Intercept|visibilityPoor|lightNight|rdsfcondPoor|road_classMajor|road_classMinor|acclocOther)") +
-          ggtitle("Trace Plots for Injury Occurrence Model - Part 1")
-```
-
-```{r, echo=FALSE, message=FALSE, warning=FALSE}
-#| label: fig-trace_o2
-#| fig-cap: "Trace Plots for middle 7 parameters in Injury Occurrence Model"
-# Next 7 parameters
-mcmc_trace(as_draws(occurrence_model), 
-          regex_pars = "b_(acclocPrivate|traffctlOther|traffctlSignal|traffctlStop|impactypeApproaching|impactypeCyclistCollisions|impactypeOther)") +
-          ggtitle("Trace Plots for Injury Occurrence Model - Part 2")
-```
-
-```{r, echo=FALSE, message=FALSE, warning=FALSE}
-#| label: fig-trace_o3
-#| fig-cap: "Trace Plots for last 6 parameters in Injury Occurrence Model"
-# Last 8 parameters
-mcmc_trace(as_draws(occurrence_model), 
-          regex_pars = "b_(impactypePedestrianCollisions|impactypeRearEnd|impactypeSideswipe|impactypeSMVOther|impactypeSMVUnattendedVehicle|impactypeTurningMovement)") +
-          ggtitle("Trace Plots for Injury Occurrence Model - Part 3")
-```
-
-@fig-trace_o1, @fig-trace_o2, @fig-trace_o3 The trace plots display stable mixing patterns across all model parameters over iterations.
-The four MCMC chains, shown in different blue shades, consistently overlap and explore similar value ranges, indicating good convergence.
-The stable patterns and lack of trending in the sampling trajectories suggest reliable parameter estimation.
-Environmental factors show relatively narrow ranges, while impact type coefficients exhibit wider ranges, reflecting their relative importance in predicting injury occurrence.
-
-\newpage
-
-```{r, echo=FALSE, message=FALSE, warning=FALSE}
-#| label: fig-trace_s1
-#| fig-cap: "Trace Plots for first six parameters in Injury Secerity Model"
-# First plot - first group of parameters
-mcmc_trace(as_draws(severity_model), 
-          regex_pars = "b_(Intercept|speeding|ag_driv|alcohol|disability|age)") +
-          ggtitle("Trace Plots for Injury Severity Model - Part 1")
-```
-
-```{r, echo=FALSE, message=FALSE, warning=FALSE}
-#| label: fig-trace_s2
-#| fig-cap: "Trace Plots for rest parameters in Injury Secerity Model"
-# Second plot - second group of parameters
-mcmc_trace(as_draws(severity_model), 
-          regex_pars = "b_(impact|light|road)") +
-          ggtitle("Trace Plots for Injury Severity Model - Part 2")
-```
-@fig-trace_s1, @fig-trace_s2 The trace plots for various model parameters display stable MCMC sampling convergence across all four chains (represented in different shades of blue).
-Each parameter's trace shows consistent exploration within its respective range without any concerning patterns or trends, indicating good mixing of the chains.
-The overlapping nature of the chains and their stationary behavior throughout the sampling period provides strong evidence that the model has reached convergence and the posterior distributions are reliable estimates of the true parameter values.
-
-\newpage
 
@@
 ## Model result
-Our dual-model analysis of traffic injuries reveals distinct patterns in both injury occurrence and severity prediction, with both models demonstrating strong predictive performance across different scenarios and conditions.
-Key findings from our models reveal the relative importance of different predictors in determining both injury occurrence and severity, and we summarise it in the follow coefficients tables and ROC curve plot.
+Our dual model analysis of traffic injuries demonstrates distinct patterns in both injury occurrence and severity prediction, key findings from our models show the relative importance of different predictors in determining both injury occurrence and severity, and we summarise it in the follow tables and ROC curve plot.
 
 ```{r, echo=FALSE, message=FALSE, warning=FALSE}
-#| label: tbl-occurrence
-#| tbl-cap: "Occurrence Model Coefficients"
-# For occurrence model
-occurrence_summary <- as.data.frame(summary(occurrence_model)$fixed)
-occurrence_table <- occurrence_summary %>%
-  rownames_to_column("Variable") %>%
-  select(Variable, Estimate, `Est.Error`, `l-95% CI`, `u-95% CI`, `Bulk_ESS`, `Tail_ESS`) %>%
-  rename(
-    'Std. Error' = 'Est.Error',
-    'Lower CI' = 'l-95% CI',
-    'Upper CI' = 'u-95% CI',
-    'Bulk ESS' = 'Bulk_ESS',
-    'Tail ESS' = 'Tail_ESS'
+#| label: tbl-summary_modeloccurrence
+#| tbl-cap: "Summary for Occurrence model"
+modelsummary(
+  occurrence_model,
+  output = "kableExtra",
 ) 
-
-# Create formatted tables
-kable(occurrence_table,
-      format = "pipe",
-      digits = 3)
 ```
 
-@tbl-occurrence The Occurrence Model uncovers several noteworthy patterns in predicting the likelihood of injuries in traffic collisions.
+@tbl-summary_modeloccurrence The Occurrence Model uncovers several noteworthy patterns in predicting the likelihood of injuries in traffic collisions.
 Among the environmental factors, nighttime conditions (0.131) are associated with a higher likelihood of injury, highlighting the risks associated with reduced visibility and nighttime driving.
@@
 However, the explanation may lie in the polarizing nature of these incidents many result in no injury at all or extreme injuries, which could skew their average likelihood of injury occurrence.
+
 Most infrastructure-related variables, such as road classification and traffic control measures, show relatively modest or non-significant effects, with confidence intervals crossing zero.
@@
 
+\newpage
+
 ```{r, echo=FALSE, message=FALSE, warning=FALSE}
-#| label: tbl-severity
-#| tbl-cap: "Severity Model Coefficients"
-# For severity model
-severity_summary <- as.data.frame(summary(severity_model)$fixed)
-severity_table <- severity_summary %>%
-  rownames_to_column("Variable") %>%
-  select(Variable, Estimate, `Est.Error`, `l-95% CI`, `u-95% CI`, `Bulk_ESS`, `Tail_ESS`) %>%
-  rename(
-    'Std. Error' = 'Est.Error',
-    'Lower CI' = 'l-95% CI',
-    'Upper CI' = 'u-95% CI',
-    'Bulk ESS' = 'Bulk_ESS',
-    'Tail ESS' = 'Tail_ESS'
+#| label: tbl-summary_modelseverity
+#| tbl-cap: "Summary for Severity model"
+modelsummary(
+  severity_model,
+  output = "kableExtra",
 ) 
-
-cat("\nTable 5: Coefficients for Severity Model\n")
-kable(severity_table,
-      format = "pipe",
-      digits = 3)
 ```
 
-@tbl-severity The results of the Severity Model reveal important insights into the factors contributing to severe injuries in motor vehicle collisions.
+@tbl-summary_modelseverity The model analysis is based on 4,572 cases that injury happened in collision, with an R-squared value of 0.111 indicating that about 11.1% of the variance in injury severity is explained by the included predictors.
 Collisions involving cyclists (2.31) and pedestrians (1.73) stand out as the most significant predictors of severe outcomes, emphasizing the heightened vulnerability of these road users.
@@
 On the other hand, factors like speeding and alcohol involvement, while often associated with risky driving behaviors, did not show a strong connection to injury severity in this dataset.
+
+\newpage
 
@@
 The curve rises sharply from the origin and maintains considerable distance from the diagonal reference line (which represents random guess), indicating strong predictive power.
-The smooth progression of the curve suggests stable performance across various sensitivity-specificity trade-offs, while its position well above the diagonal line confirms the model's robust ability to distinguish between injury severity classes.
+The smooth progression of the curve suggests stable performance across various sensitivity-specificity trade offs, while its position well above the diagonal line confirms the model's robust ability to distinguish between injury severity classes.
 
 ## Prediction example
-Our prediction analysis utilizes scenario-based datasets (occurrence_scenarios and severity_scenarios) constructed with expand_grid to systematically explore various combinations of environmental, infrastructural, and behavioral conditions.
-By inputting these scenarios into our models, we generate predicted probabilities of injury occurrence and severity across a comprehensive range of real-world situations.
+Our prediction analysis utilizes scenario-based datasets (occurrence_scenarios and severity_scenarios) to systematically explore various combinations of environmental, infrastructural, and behavioral conditions.
+By inputting these scenarios into our models, we generate predicted probabilities of injury occurrence and severity across a rich range of real world situations.
 
@@
 
-@fig-avp The actual versus predicted probabilities plot for the injury severity model reveals a strong predictive performance, with observations clustering predominantly at either high (around 1.0) or low (around 0.0) probabilities.
+@fig-avp The actual versus predicted probabilities plot for the injury severity model shows a strong predictive performance, with observations clustering predominantly at either high (around 1.0) or low (around 0.0) probabilities.
 The blue diagonal line represents perfect prediction, and the pattern of actual outcomes shows clear separation between severity classes, indicating the model's ability to effectively discriminate between minor and severe injuries.
-This binary clustering pattern is characteristic of a well-calibrated binary classification model.
+This binary clustering pattern is characteristic of a good calibrated binary classification model.
 
@@
 # Discussion {#sec-discussion}
+## Potential real world improvement based on result
+Given the heightened risks associated with nighttime conditions and single-vehicle incidents, municipalities should prioritize infrastructure improvements that specifically address these vulnerabilities.
+This could include implementing street lighting systems in high-risk areas, installing reflective road markers and enhanced signage on curves and potentially hazardous stretches, and deploying smart traffic monitoring systems that adjust speed limits based on visibility conditions.
+For minor roads, where our analysis shows a concerning pattern of severe injuries, some safety upgrades should include traffic calming measures such as speed humps, raised crosswalks, and chicanes.
+These improvements, combined with automated enforcement systems, could help mitigate the risks that make these seemingly less-traveled roads particularly dangerous.
+
+The findings also empasis the need for a reformed approach to protecting vulnerable road users and emergency response protocols.
+Given the high severity coefficients associated with cyclist and pedestrian incidents, cities should invest in protected infrastructure networks that physically separate these vulnerable users from motor vehicle traffic, particularly along high-traffic corridors and near community hubs.
+Emergency services can enhance their response effectiveness by implementing a tiered dispatch system that automatically elevates the response priority for incidents involving cyclists, pedestrians, and senior citizens - groups our analysis identifies as being at higher risk for severe injuries.
+
 ## Data quality and limitations
@@
 The primary collision dataset from opendatatoronto [@opendatatoronto] contains inherent biases in injury reporting.
-According to the National Highway Traffic Safety Administration's recent report [@nhtsa2023], up to 40% of minor injury accidents go unreported, particularly in cases involving single vehicles or where parties choose to handle incidents privately.
+According to the National Highway Traffic Safety Administration's recent report [@nhtsa2023], lots of case about minor injury accidents go unreported, particularly in cases involving single vehicles or where parties choose to handle incidents privately.
 This underreporting likely creates a selection bias towards more severe accidents in our dataset, potentially affecting our model's predictions for minor injury scenarios.
@@
 Environmental conditions are simplified into binary categories (e.g., "Good"/"Poor" visibility, "Day"/"Night" lighting) to maintain model stability and interpretability.
-However, this simplification may mask important nuances in intermediate conditions.
+However, this simplification may mask important differences in intermediate conditions.
 For example, the effect of partial visibility or dusk conditions might differ significantly from either clear or poor visibility, but our current categorization cannot capture these distinctions.
@@
 Several promising directions for improvement emerge from our analysis.
-First, the integration of continuous monitoring data could provide more granular insights into accident patterns.
-This could include real-time traffic flow data, weather conditions, and road surface measurements, allowing for more nuanced understanding of how environmental conditions affect accident risk.
+First, the integration of continuous monitoring data could provide more granular understand for accident patterns.
+This could include real-time traffic flow data, weather conditions, and road surface measurements, allowing for more precise understanding of how environmental conditions affect accident risk.
 
@@
 # Appendix {-}
-## Data collecting methodology {#sec-data-methodology}
+# Data collecting methodology {#sec-data-methodology}
 The City of Toronto's Motor Vehicle Collision database involving Killed or Seriously Injured Persons (KSI) is collected through a systematic process involving the Toronto Police Service (TPS) as the primary data collector [@toronto_collision_data].
@@
 
-A critical aspect of the data collection methodology is its focus on Killed or Seriously Injured (KSI) events.
+A noticeable aspect of the data collection methodology is its focus on Killed or Seriously Injured (KSI) events.
 The dataset defines major injury as "a non-fatal injury severe enough to require the injured person to be admitted to hospital, even if only for observation." This includes fractures, internal injuries, severe cuts, crushing, burns, concussion, and severe general shocks.
@@
 
-The dataset's structure reveals a careful attention to detail in recording various aspects of each collision.
+The dataset's structure shows a careful attention to detail in recording various aspects of each collision.
 Each record represents a person involved in a KSI collision event, regardless of their injury level.
@@
 
-A significant methodological strength is the comprehensive categorization system used for various collision attributes.
+A significant methodological strength is the categorization system used for various collision attributes.
 The dataset includes detailed classifications for environmental conditions, road surface conditions, collision types, and participant behaviors [@toronto_police_services_ksi_guide].
@@
 
-## Idealized methodology {#sec-data-idealized}
+# Idealized methodology {#sec-data-idealized}
 An idealized methodology for motor vehicle collision data collection must fundamentally reimagine how we gather, validate, and integrate traffic incident information while maintaining robust privacy protections.
-Drawing from advancements in digital technology and data science, I propose a comprehensive modernization of the current system.
+Drawing from advancements in digital technology and data science, we propose a modernization for the current system.
 The cornerstone of this enhanced methodology would be transitioning from paper-based documentation to a sophisticated digital data collection framework.
 Police officers would utilize specialized mobile devices equipped with purpose-built software that guides them through systematic data collection at collision scenes [@digital_transformation].
-This digital transformation would significantly reduce transcription errors while enabling real-time data validation - a crucial improvement over the current SR-LD-401 form system.
+This digital transformation would significantly reduce transcription errors while enabling real time data validation, a big improvement over the current SR-LD-401 form system.
 
@@
 The current practice of offsetting collision locations to the nearest intersection, while serving its primary purpose, introduces unnecessary spatial imprecision.
-Advanced geographic anonymization techniques could better balance privacy concerns with analytical precision.
+Some geographic anonymizations techniques could better balance privacy concerns with analytical precision.
 For instance, implementing differential privacy algorithms could add calibrated noise to geographic coordinates while preserving statistical accuracy for analysis purposes [@geographic_anonymization].
@@
 Environmental context represents another area where current methodologies rely heavily on subjective officer assessment.
-A modernized system would automatically integrate data from multiple validated sources, creating a more objective and comprehensive picture of collision conditions.
-This would include real-time weather data from official weather stations, traffic volume information from road sensors, and current road condition data from municipal databases.
+A modernized system would automatically integrate data from multiple validated sources, creating a more objective understand of collision conditions.
+This would include real time weather data from official weather stations, traffic volume information from road sensors, and current road condition data from municipal databases.
 Such integration would provide a more reliable environmental context for each incident.
@@
 By cross referencing police reports with emergency medical services records, hospital admission data, and insurance claims, while employing privacy-preserving record linkage techniques and the system would capture a more complete picture of collision incidents and their outcomes.
+
 This idealized methodology may require significant investment in technology infrastructure and training, but it represents a necessary evolution in traffic collision data collection, which resulting improvements in data quality and data completeness.
 
-## Model details {#sec-model-details}
+# Survey of collision patterns recognization in Toronto's residents 
+To gather additional information about vehicle collision patterns in Toronto based on residents, we propose a survey targeted at residents in Toronto.
+This survey aims to collect data about personal experiences with traffic conditions, observed patterns of dangerous driving, and environmental factors affecting road safety.
+Below is a survey to collect this information.
 
+What is your age group?
+
+- 16-24
+
+- 25-34
+
+- 35-49
+
+- 50-64
+
+- 65 and above
+
+What is your gender?
+
+- Male
+
+- Female
+
+- Other
+
+- Prefer not to say
+
+In which area of Toronto do you primarily drive? (Select all that apply)
+
+- Downtown Core
+
+- North York
+
+- Scarborough
+
+- Etobicoke
+
+How frequently do you drive in Toronto?
+
+- Daily
+
+- Several times per week
+
+- Weekly
+
+- Monthly
+
+- Rarely
+
+In your experience, which time periods have the highest collision risk?
+
+- Morning rush hour (7-9 AM)
+
+- Evening rush hour (4-6 PM)
+
+- Late night/early morning
+
+- Weekend afternoons
+
+Which environmental factors do you think most affect driving safety? (Select all that apply)
+
+- Poor visibility
+
+- Wet road conditions
+
+- Snow/ice
+
+- Construction
+
+- Heavy traffic
+
+Based on your experience, which road types seem most prone to collisions?
+
+- Major arterial roads
+
+- Local residential streets
+
+- Highway ramps
+
+- Intersections
+
+Have you noticed any season have more collision happen?
+
+- More incidents in winter
+
+- More incidents in summer
+
+- No seasonal pattern
+
+- Not sure
+
+Which driver behaviors do you most commonly observe that increase collision risk? (Select all that apply)
+
+- Speeding
+
+- Aggressive driving
+
+- Distracted driving
+
+- Following too closely
+
+- Improper turns
+
+- Other
+
+In your opinion, how effective are current traffic control measures (signals, signs, speed limits) in preventing collisions?
+
+- Very effective
+
+- Somewhat effective
+
+- Neutral
+
+- Somewhat ineffective
+
+- Very ineffective
+
+Have you been involved in or witnessed a collision in Toronto in the past 5 years?
+
+- Yes, involved in a collision
+
+- Yes, witnessed a collision
+
+- No
+
+- Prefer not to say
+
+If you answered yes to the previous question, what factors contributed to the collision? (Select all that apply)
+
+- Weather conditions
+
+- Road conditions
+
+- Driver behavior
+
+- Traffic control issues
+
+- Vehicle issues
+
+- Other
+
+# Model validations and details {#sec-model-validate_and_details}
+## Model validation {#sec-model-validate}
+Our validation approach employs a sophisticated stratified sampling methodology, dividing the data into training (70%) and testing (30%) sets while carefully maintaining the original class proportions.
+This strategy is significant given the inherently imbalanced nature of traffic injury data.
+The validation framework encompasses multiple complementary approaches, including confusion matrix analysis for understanding classification performance, receiver operating characteristic (ROC) curve analysis for assessing discriminative ability.
+We further ensure model stability through cross validation procedures.
+Prior to model fitting, we conducted thorough multicollinearity assessments using Variance Inflation Factors, confirming the independence of our predictor variables.
+
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
+#| label: fig-posterior
+#| fig-cap: "Posterior Predictive Check for Injury Occurrence Model"
+pp_check(occurrence_model, type = "bars", ndraws = 100) +
+  ggtitle("Posterior Predictive Check for Injury Occurrence Model")
+```
+@fig-posterior The posterior predictive check plot demonstrates the model's ability to replicate the binary nature of injury occurrence data.
+The blue dots (yrep) represent simulated data from the posterior predictive distribution, while the light blue bars (y) show the actual observed data distribution.
+The plot shows two clear peaks around 0 and 1, corresponding to no-injury and injury cases respectively, with the simulated data closely matching the observed distribution.
+The alignment between predicted and actual counts at both peaks suggests that the model effectively captures the underlying binary structure of injury occurrences while maintaining appropriate uncertainty in its predictions.
+
+\newpage
+
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
+#| label: fig-trace_o1
+#| fig-cap: "Trace Plots for first 7 parameters in Injury Occurrence Model"
+# First 7 parameters
+mcmc_trace(as_draws(occurrence_model), 
+          regex_pars = "b_(Intercept|visibilityPoor|lightNight|rdsfcondPoor|road_classMajor|road_classMinor|acclocOther)") +
+          ggtitle("Trace Plots for Injury Occurrence Model - Part 1") +
+  theme(axis.text.x = element_text(size = 7),
+        axis.text.y = element_text(size = 7),
+        axis.title.x = element_text(size = 9),
+        axis.title.y = element_text(size = 9)
+)
+```
+
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
+#| label: fig-trace_o2
+#| fig-cap: "Trace Plots for middle 7 parameters in Injury Occurrence Model"
+# Next 7 parameters
+mcmc_trace(as_draws(occurrence_model), 
+          regex_pars = "b_(acclocPrivate|traffctlOther|traffctlSignal|traffctlStop|impactypeApproaching|impactypeCyclistCollisions|impactypeOther)") +
+          ggtitle("Trace Plots for Injury Occurrence Model - Part 2") +
+  theme(axis.text.x = element_text(size = 7),
+        axis.text.y = element_text(size = 7),
+        axis.title.x = element_text(size = 9),
+        axis.title.y = element_text(size = 9)
+)
+```
+
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
+#| label: fig-trace_o3
+#| fig-cap: "Trace Plots for last 6 parameters in Injury Occurrence Model"
+# Last 8 parameters
+mcmc_trace(as_draws(occurrence_model), 
+          regex_pars = "b_(impactypePedestrianCollisions|impactypeRearEnd|impactypeSideswipe|impactypeSMVOther|impactypeSMVUnattendedVehicle|impactypeTurningMovement)") +
+          ggtitle("Trace Plots for Injury Occurrence Model - Part 3") +
+  theme(axis.text.x = element_text(size = 7),
+        axis.text.y = element_text(size = 7),
+        axis.title.x = element_text(size = 9),
+        axis.title.y = element_text(size = 9)
+)
+```
+
+@fig-trace_o1, @fig-trace_o2, @fig-trace_o3 The trace plots display stable mixing patterns across all model parameters over iterations.
+The four MCMC chains, shown in different blue shades, consistently overlap and explore similar value ranges, indicating good convergence.
+The stable patterns and lack of trending in the sampling trajectories suggest reliable parameter estimation.
+Environmental factors show relatively narrow ranges, while impact type coefficients exhibit wider ranges, reflecting their relative importance in predicting injury occurrence.
+
+\newpage
+
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
+#| label: fig-trace_s1
+#| fig-cap: "Trace Plots for first six parameters in Injury Secerity Model"
+# First plot - first group of parameters
+mcmc_trace(as_draws(severity_model), 
+          regex_pars = "b_(Intercept|speeding|ag_driv|alcohol|disability|age)") +
+          ggtitle("Trace Plots for Injury Severity Model - Part 1") +
+  theme(axis.text.x = element_text(size = 7),
+        axis.text.y = element_text(size = 7),
+        axis.title.x = element_text(size = 9),
+        axis.title.y = element_text(size = 9)
+)
+```
+
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
+#| label: fig-trace_s2
+#| fig-cap: "Trace Plots for rest parameters in Injury Secerity Model"
+# Second plot - second group of parameters
+mcmc_trace(as_draws(severity_model), 
+          regex_pars = "b_(impact|light|road)") +
+          ggtitle("Trace Plots for Injury Severity Model - Part 2") +
+  theme(axis.text.x = element_text(size = 7),
+        axis.text.y = element_text(size = 7),
+        axis.title.x = element_text(size = 9),
+        axis.title.y = element_text(size = 9)
+)
+```
+@fig-trace_s1, @fig-trace_s2 The trace plots for various model parameters display stable MCMC sampling convergence across all four chains (represented in different shades of blue).
+Each parameter's trace shows consistent within its respective range without any concerning patterns or trends, indicating good mixing of the chains.
+The overlapping nature of the chains and their stationary behavior throughout the sampling period provides strong evidence that the model has reached convergence and the posterior distributions are reliable estimates of the true parameter values.
+
+\newpage
+
+## Model details {#sec-model-detail}
 ```{r, echo=FALSE, message=FALSE, warning=FALSE}
@@
 kable(fixed_effects, 
-      format = "markdown",  # Choose the desired format ("markdown", "html", or "latex")
-      digits = 3,           # Round numbers to 3 decimal places
-      col.names = c("Estimate", "Est. Error", "l-95% CI", "u-95% CI", "Rhat", "Bulk ESS", "Tail ESS"), # Rename columns
-      caption = "Summary of Fixed Effects for the Model")
+      format = "markdown",
+      digits = 3,
+      col.names = c("Estimate", "Est. Error", "l-95% CI", "u-95% CI", "Rhat", "Bulk ESS", "Tail ESS")
+)
 ```
+
+\newpage
 
@@
 kable(fixed_effects, 
-      format = "markdown",  # Choose the desired format ("markdown", "html", or "latex")
-      digits = 3,           # Round numbers to 3 decimal places
-      col.names = c("Estimate", "Est. Error", "l-95% CI", "u-95% CI", "Rhat", "Bulk ESS", "Tail ESS"), # Rename columns
-      caption = "Summary of Fixed Effects for the Model")
+      format = "markdown",
+      digits = 3,
+      col.names = c("Estimate", "Est. Error", "l-95% CI", "u-95% CI", "Rhat", "Bulk ESS", "Tail ESS")
+)
 ```
 
-\newpage
+\clearpage
 

### 2024-12-04 paper/paper.qmd
Refine the title and readme

@@
 ---
-title: "Environmental Conditions and Driver Behavior Affect Collision Outcomes in Toronto"
+title: "Motor Traffic Collision Patterns in Toronto: Analysis based on Environmental Conditions and Driver Behaviours"
 subtitle: "A Two-Stage Bayesian Analysis Shows Infrastructure Factors Predict Crash Occurrence While Driver Actions Determine Severity"
@@
 Analysis of historical collision data from 2006 to 2023 shows that while environmental conditions primarily influence injury occurrence, driver behavior factors like speeding and alcohol use more strongly affect severity levels.
-These findings demonstrate the need for a dual-pronged approach to urban safety through targeted infrastructure improvements in high-risk areas to prevent collisions, alongside enhanced enforcement strategies focusing on dangerous driving behaviors.
-"
+These findings demonstrate the need for a dual-pronged approach to urban safety through targeted infrastructure improvements in high-risk areas to prevent collisions, alongside enhanced enforcement strategies focusing on dangerous driving behaviors."
 format: pdf

### 2024-12-04 scripts/03-clean_data.R
Update R package

@@ -13,6 +13,7 @@
 library(tidyverse)
 library(here)
 library(arrow)
+library(janitor)
 
 #### Clean data ####
 # Read raw data

### 2024-12-14 other/datasheet/collision_datasheet.qmd
Update collision datasheet and relevant files

---
title: "Datasheet for 'Motor Vehicle Collisions in Toronto'"
subtitle: "Motivation, Composition, Collection, Preprocessing, Uses, Distribution, and Maintenence of the Dataset"
author: Yitong Wang
thanks: "Code and data are available at: [https://github.com/demainwang/Toronto_vehicle_collision_analysis](https://github.com/demainwang/Toronto_vehicle_collision_analysis)."
date: today
date-format: long
abstract: "The Toronto Motor Vehicle Collisions dataset provides documentation of traffic incidents resulting in fatalities or serious injuries within the city from 2006 to present. Police department collects detailed information about collision locations, environmental conditions, and involved parties through standardized reporting procedures. Each record captures essential details including geographic coordinates, time information, road conditions, and injury severity classifications, while carefully balancing public safety interests with privacy protection by excluding personal identifiers. This dataset serves multiple stakeholders including urban planners, researchers, and community organizations, supporting evidence based decision making for road safety improvements."
format: pdf
number-sections: true
bibliography: references.bib
---

Extract of the questions from [@gebru2021datasheets].
The Motor Vehicle Collisions involving Killed or Seriously Injured Persons dataset [@toronto_collision_data] in this discussion and study is obtained from opendatatoronto [@opendatatoronto]. Some background information comes from Killed or Seriously Injured Glossary [@toronto_police_services_ksi_guide] and Motor Vehicle Collision Report Guide [@mvcr_guide].

**Motivation**

1. *For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*
    - The dataset was created to enable analysis of serious and fatal motor vehicle collisions in Toronto. The dataset fills an gap in providing transparent, detailed information about severe traffic incidents, it combines both environmental factors and driver factors to understand the pattern of injury occurrence and severity to help inform infrastructure improvements, and ultimately prevent future collisions. This data collection effort aligns with Toronto's broader commitment to improving road safety and public transparency.

2. *Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*
    - The dataset was created by the city of Toronto's transportation services division in collaboration with Toronto police service, who are responsible for collecting and documenting collision data at the scene and through subsequent investigations.
3. *Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*
    - No specific grantor fund this dataset, this dataset based on the record of collision cases.

**Composition**

1. *What do the instances that comprise the dataset represent (for example, documents, photos, people, countries)? Are there multiple types of instances (for example, movies, users, and ratings; people and interactions between them; nodes and edges)? Please provide a description.*
	- Each instance documents a motor vehicle collision in Toronto from 2006 to 2023 where someone was killed or seriously injured. The records include environmental conditions, location details, and circumstantial factors for each incident.
2. *How many instances are there in total (of each type, if appropriate)?*
	- The dataset includes 14,976 records, each representing an individual involved in a collision.
3. *Does the dataset contain all possible instances or is it a sample (not necessarily random) of instances from a larger set? If the dataset is a sample, then what is the larger set? Is the sample representative of the larger set (for example, geographic coverage)? If so, please describe how this representativeness was validated/verified. If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).*
	- The dataset aims to contain all reported serious injury and fatal collisions that occurred within Toronto city and were reported to Toronto Police Service. It can be considered as a subset of all motor traffic collision happen in Toronto, but this dataset focus on those case involve human injury.
4. *What data does each instance consist of? "Raw" data (for example, unprocessed text or images) or features? In either case, please provide a description.*
	- Each record contains event specifics (date, time, location), environmental factors (weather, road conditions), collision details (type, impact), geographic coordinates such as latitude or longitude, involved driver related information (age group, driving conditions), injury severity, road type, and traffic control presence.
5. *Is there a label or target associated with each instance? If so, please provide a description.*
	- Records are labeled by outcome severity (level of injury).
6. *Is any information missing from individual instances? If so, please provide a description, explaining why this information is missing (for example, because it was unavailable). This does not include intentionally removed information, but might include, for example, redacted text.*
	-  All records are not include the exact speed of traffic when incident happened. Some records have incomplete fields, particularly in older cases. Weather conditions occasionally show missing values when officers could not definitively determine conditions at the time of reporting. Personal identifying information is omitted to protect privacy, including names, exact ages. Exact incident location have been deliberately offset to the nearest road intersection node to protect the privacy.
7. *Are relationships between individual instances made explicit (for example, users' movie ratings, social network links)? If so, please describe how these relationships are made explicit.*
	- The dataset does not explicitly link separate incidents even if they might share common characteristics. Location based relationships may link incidents occurring at the same intersection or road segment. Temporal relationships may connect collisions occurring during similar conditions or time periods.
8. *Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.*
	- The data naturally divides into temporal splits, with recommendations to separate analysis by year to account for changes in reporting practices and road safety measures. The dataset can be split geographically by district or ward for local analysis. However, no official training and testing splits exist as this represents real world incident data rather than a machine learning dataset.
9. *Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.*
	- Initial reporting might include estimation errors in environmental conditions or exact collision timing, the recogniztion of driver conditions are highly rely on officer's subjective thinking, might be bias.
10. *Is the dataset self-contained, or does it link to or otherwise rely on external resources (for example, websites, tweets, other datasets)? If it links to or relies on external resources, a) are there guarantees that they will exist, and remain constant, over time; b) are there official archival versions of the complete dataset (that is, including the external resources as they existed at the time the dataset was created); c) are there any restrictions (for example, licenses, fees) associated with any of the external resources that might apply to a dataset consumer? Please provide descriptions of all external resources and any restrictions associated with them, as well as links or other access points, as appropriate.*
	- The dataset connects to several external systems while maintaining its core integrity. It links to the City of Toronto's geographic information system for location validation and mapping. The dataset references standardized coding systems for collision types and injury classifications used by the Toronto Police Service.
11. *Does the dataset contain data that might be considered confidential (for example, data that is protected by legal privilege or by doctor-patient confidentiality, data that includes the content of individuals' non-public communications)? If so, please provide a description.*
	- The dataset contains sensitive information about traffic collisions involving deaths and serious injuries. While direct personal identifiers are removed, the combination of location, time, and incident details could potentially allow identification of individuals involved in high-profile cases. Medical information appears only in broad categories of injury severity rather than specific details.
12. *Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.*
	- The dataset documents serious and fatal traffic collisions, which could cause distress to readers, particularly those who have experienced similar incidents or lost loved ones in traffic collisions. The descriptions remain factual and technical, focusing on circumstances.
13. *Does the dataset identify any sub-populations (for example, by age, gender)? If so, please describe how these subpopulations are identified and provide a description of their respective distributions within the dataset.*
	- It includes basic demographic information but avoids detailed personal characteristics. The data allows analysis of collision patterns affecting different road users while maintaining individual privacy. Geographic distribution of incidents might correlate with neighborhood demographics.
14. *Is it possible to identify individuals (that is, one or more natural persons), either directly or indirectly (that is, in combination with other data) from the dataset? If so, please describe how.*
	- While the dataset removes direct personal identifiers, the combination of location, date, time, and incident characteristics might enable indirect identification in some cases, particularly for notable incidents that received media coverage.
15. *Does the dataset contain data that might be considered sensitive in any way (for example, data that reveals race or ethnic origins, sexual orientations, religious beliefs, political opinions or union memberships, or locations; financial or health data; biometric or genetic data; forms of government identification, such as social security numbers; criminal history)? If so, please provide a description.*
	- The dataset contains several categories of sensitive information. Location data shows where serious incidents occur, which could affect neighborhood perceptions. The timing and circumstances of fatal collisions represent sensitive details for affected families and communities. The data includes broad age groups of those involved, though it excludes other demographic details to protect privacy.

**Collection process**

1. *How was the data associated with each instance acquired? Was the data directly observable (for example, raw text, movie ratings), reported by subjects (for example, survey responses), or indirectly inferred/derived from other data (for example, part-of-speech tags, model-based guesses for age or language)? If the data was reported by subjects or indirectly inferred/derived from other data, was the data validated/verified? If so, please describe how.*
	- Toronto Police Service officers collect data directly at collision scenes through standardized reporting procedures. Officers observe the scene, take measurements, and document environmental conditions.
2. *What mechanisms or procedures were used to collect the data (for example, hardware apparatuses or sensors, manual human curation, software programs, software APIs)? How were these mechanisms or procedures validated?*
	- Police officers use electronic reporting systems installed in their vehicles to document collision details on-site. They follow standardized protocols established by the Toronto Police Service for collision investigation and reporting. 
3. *If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
	- This dataset represents the complete record of collisions involving fatalities or serious injuries in Toronto. Every qualifying incident undergoes documentation and inclusion in the dataset. 
4. *Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
	- Toronto Police Service officers serve as primary data collectors as part of their regular duties. All personnel involved receive compensation through their standard government employment arrangements.
5. *Over what timeframe was the data collected? Does this timeframe match the creation timeframe of the data associated with the instances (for example, recent crawl of old news articles)? If not, please describe the timeframe in which the data associated with the instances was created.*
	- The dataset covers collisions from 2006 to present, with continuous updates as new incidents occur. The collection timeframe matches actual collision occurrences.
6. *Were any ethical review processes conducted (for example, by an institutional review board)? If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- The data collection operates under the Toronto Police Service's established protocols for incident reporting, The public release process includes privacy impact assessments to protect individual rights, including hide exact location, names and age.
7. *Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*
	- The Toronto Police Service obtains information directly through officer investigations at collision scenes. In cases of serious injuries, medical authorities provide injury classification updates while maintaining patient privacy. The dataset represents official police records rather than third party collected information.
8. *Were the individuals in question notified about the data collection? If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.*
	- When police officers investigate collisions, they inform involved parties about official reporting requirements and data collection. The public nature of serious collision investigation means participants understand that basic incident information becomes part of public safety records, though personal details remain protected.
9. *Did the individuals in question consent to the collection and use of their data? If so, please describe (or show with screenshots or other information) how consent was requested and provided, and provide a link or other access point to, or otherwise reproduce, the exact language to which the individuals consented.*
	- The data collection occurs under the legal authority of the Toronto Police Service to investigate and document serious collisions. The public release version removes personal identifiers to protect privacy while maintaining the public safety benefits of collision data analysis.
10. *If consent was obtained, were the consenting individuals provided with a mechanism to revoke their consent in the future or for certain uses? If so, please provide a description, as well as a link or other access point to the mechanism (if appropriate).*
	- Given the public safety mandate for collision reporting, the dataset does not revocate related information. 
11. *Has an analysis of the potential impact of the dataset and its use on data subjects (for example, a data protection impact analysis) been conducted? If so, please provide a description of this analysis, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- There was not official analysis of potential impact.

**Preprocessing/cleaning/labeling**

1. *Was any preprocessing/cleaning/labeling of the data done (for example, discretization or bucketing, tokenization, part-of-speech tagging, SIFT feature extraction, removal of instances, processing of missing values)? If so, please provide a description. If not, you may skip the remaining questions in this section.*
	- Not specific mention of details of data processing, but it made change for location data, which deliberately offset to the nearest road intersection node to protect the privacy.
2. *Was the "raw" data saved in addition to the preprocessed/cleaned/labeled data (for example, to support unanticipated future uses)? If so, please provide a link or other access point to the "raw" data.*
	- The original raw data not published due to the sensitive information about people's privacy.
3. *Is the software that was used to preprocess/clean/label the data available? If so, please provide a link or other access point.*
	- Not specific mention about the software, but generally it may use the police internal system to handle the data.

**Uses**

1. *Has the dataset been used for any tasks already? If so, please provide a description.*
	- Not specific mention of the previous using of data, but transportation services may uses this dataset to identify high risk locations requiring safety improvements. City planners incorporate the data into Vision Zero safety initiatives. 
2. *Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*
	- Have not found it.
3. *What (other) tasks could the dataset be used for?*
	- The data could support urban planning decisions, emergency response resource allocation, and public education about road safety.
4. *Is there anything about the composition of the dataset or the way it was collected and preprocessed/cleaned/labeled that might impact future uses? For example, is there anything that a dataset consumer might need to know to avoid uses that could result in unfair treatment of individuals or groups (for example, stereotyping, quality of service issues) or other risks or harms (for example, legal risks, financial harms)? If so, please provide a description. Is there anything a dataset consumer could do to mitigate these risks or harms?*
	- The collection of this dataset mostly based on fact, but as mentioned some recording like driver conditions can have officer's subjective bias.
5. *Are there tasks for which the dataset should not be used? If so, please provide a description.*
	- The dataset should not serve for individual incident liability determinations or personal identification purposes.

**Distribution**

1. *Will the dataset be distributed to third parties outside of the entity (for example, company, institution, organization) on behalf of which the dataset was created? If so, please provide a description.*
	- The City of Toronto makes the dataset publicly available through its open Data portal, enabling access by researchers, planners, and community organizations.
2. *How will the dataset be distributed (for example, tarball on website, API, GitHub)? Does the dataset have a digital object identifier (DOI)?*
	- The Open Data Portal provides CSV and GeoJSON format downloads. Users can access the data through web interfaces and API connections.
3. *When will the dataset be distributed?*
	- The dataset updates as new verified records become available.
4. *Will the dataset be distributed under a copyright or other intellectual property (IP) license, and/or under applicable terms of use (ToU)? If so, please describe this license and/ or ToU, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms or ToU, as well as any fees associated with these restrictions.*
	- The dataset is permitting for broad application for public benefit.
5. *Have any third parties imposed IP-based or other restrictions on the data associated with the instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.*
	- This government dataset is public and not restriction by other parties.
6. *Do any export controls or other regulatory restrictions apply to the dataset or to individual instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any supporting documentation.*
	- Not such restrictions.

**Maintenance**

1. *Who will be supporting/hosting/maintaining the dataset?*
	- No specific mention, but The police and transportation Services division may maintain the dataset with technical support from the city's information technology department.
2. *How can the owner/curator/manager of the dataset be contacted (for example, email address)?*
	- Users can contact the open Data team through the portal's feedback system.
3. *Is there an erratum? If so, please provide a link or other access point.*
	- The open Data Portal maintains a change log documenting significant corrections and updates.
4. *Will the dataset be updated (for example, to correct labeling errors, add new instances, delete instances)? If so, please describe how often, by whom, and how updates will be communicated to dataset consumers (for example, mailing list, GitHub)?*
	- Updates add when new incidents are avaible and revise existing records based on finalized investigations.
5. *If the dataset relates to people, are there applicable limits on the retention of the data associated with the instances (for example, were the individuals in question told that their data would be retained for a fixed period of time and then deleted)? If so, please describe these limits and explain how they will be enforced.*
	- The dataset does not involve the exact personal information.
6. *Will older versions of the dataset continue to be supported/hosted/maintained? If so, please describe how. If not, please describe how its obsolescence will be communicated to dataset consumers.*
	- Nope, the new data will be added to the combinations of old data in one file.
7. *If others want to extend/augment/build on/contribute to the dataset, is there a mechanism for them to do so? If so, please provide a description. Will these contributions be validated/verified? If so, please describe how. If not, why not? Is there a process for communicating/distributing these contributions to dataset consumers? If so, please provide a description.*
	- Nope, this is the government dataset based on the collision information provided by police.

\newpage

# References
